### 2023-04-04 README.Rmd
Initial push

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# quartzite.complexities

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MattiasSealander/quartzite.complexities/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2023-04-04 analysis/paper/paper.qmd
Added data, updating to quarto paper

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of the same these materials. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin and potentially the metamorphic grade. These features indicate the presence of graphite, muscovite, and biotite. Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region. There also seems to be a good ground for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model could prove useful in spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

<!-- The following function can be used to suppress message outputs -->
```{r}	
quietly <- function(x) {	
  sink(tempfile())	
  on.exit(sink())	
  invisible(force(suppressMessages(x)))	
}	
```

<!-- The following code chunk is used for inline calculations of samples and sites -->
```{r sample-count, results='hide'}
suppressPackageStartupMessages(library(tidyverse))

#Import descriptive metadata
metadata.csv <-
  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")

#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
                site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
                site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
                site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
                site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
                site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
                site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
                site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
                site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
                site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")

#Create a vector containing all unique sample ID
tot_uniq_samples <- 
  unique(tot_samples$sample_id)

#Create a vector containing all unique site ID
tot_uniq_sites <-
  unique(tot_samples$site_id)

#Create a vector containing all unique sample ID made in breacciated quartz
bquartz <- unique(tot_samples %>% 
  filter(material == "Brecciated quartz"))$sample_id

#Create a vector containing all unique sample ID made in quartz
quartz <- unique(tot_samples %>% 
  filter(material == "Quartz"))$sample_id

#Create a vector containing all unique sample ID made in quartzite
quartzite <- unique(tot_samples %>% 
  filter(material == "Quartzite"))$sample_id
```

# Introduction

Petrographic characterization and provenance studies are instrumental for inferences of trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent has been highlighted by Prieto et al. [-@prieto_etal19: 15]. In Northern Sweden both quartz and quartzite dominate the lithic assemblages and have been used in studies of settlements, exchange and mobility [@baudou78]. While quartz and quartzite technology have been the target of several studies over the years, there has been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. M While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden haves been put forth [@halen94: 115-122; @lundberg97: 161-171]. On the other hand there haves been little development on the topic of quartz and quartzite [@holm91: 23-24]. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine deposits and the degree to which geological materials have been mixed and transported necessitates considerable effort to reconstruct the events that have formed the current landscape. Existing theories on lithic material sourcing within the North Swedish region tend to describe strategies regarding direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06: 249-250]. Due to this ambiguous nature of availability of raw materials, provenance studies are complex.

Prieto et al. [-@prieto_etal19: 15; -@prieto_etal20: 32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature. Because of this , and different researchers have thus emphasized different different features as being characteristics of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05: 708] also emphasises the need for field-based methodologies. Spectroscopic surface measurements are fast, and can be adapted for use in the field. In the case of archaeological research there are situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.

X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

# Aims {-}

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (@fig-distribution-map). The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.

# Material {-}

\linebreak
 \linebreak

The current study includes spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map). The sites are distributed along the Ångerman river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. No distinction is made between arrow- and spearheads.

A number of these points have, according to prior classifications, been worked using bifacial technology. The geographically closest chronology for bifacial points can be found in southeast Norway [@mjaerum12], but so far few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89], ). These attempts suggests a period of use between ca. 3 800 - 2 000 BP [@kaeck09: 60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].

\linebreak

```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."
#| fig-height: 15
#| fig-fullwidth: true

knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
```

\linebreak

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separation of quartz from quartzite based on non-destructive analysis of their chemical and structural characteristics, and the second related purpose is to account for potential mistakes in the original classification.

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91: 24]. As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

\linebreak

```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 2-76 mm. "

knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
```

\linebreak

# Methods {-}

\linebreak
 \linebreak

All measurements were performed at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens. The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape (@fig-artefacts - @fig-artefact-size) (see Supplementary Materials). The study is non-destructive sampling was performed on the surface of a material that is not uniform. The material ranges from fairly small and flat point fragments, to large and rough preforms. Therefore, the signal integration could at times be difficult, which in turn could lead to scattering effects and saturation.

Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. Sampling materials that featured high heterogeneity the sample point was focused on the surrounding matrix, away from deviating inclusions. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

The artefacts features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent. While the colour of the material has been classified according to the Munsell system, for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent). In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.

(ref:artefact)

\linebreak

```{r}
#| label: fig-artefact-size
#| cache: true
#| fig-cap: "Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. 8 objects have no recorded thickness, which ranges between 3-76 mm. Note that the term *point* does not differentiate between spearheads or arrowheads."

quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)
```

\linebreak

## Instrumentation and data pre-processing {-}

\linebreak
 \linebreak

### Near Infrared {-}
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near infrared region. The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify the bond. Overtones occur at frequencies that are a multiple of the infrared vibration. Most of them end up in the near infrared region [@osborne_etal93].

Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before PCA modelling.

\linebreak

### Raman {-}
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^. Integration time was set to 20 seconds with 30% power. After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using Lieber and Mahadevan-Jansens' [-@lieber_mahadevan-jansen03] algorithm for polynomial fitting, with the ChemoSpec package [@hanson16]. The spectra were then mean-centred before PCA.

\linebreak

### X-Ray Fluourescence {-}
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Spectral features in XRF spectrometry can be related to certain elements according to Moseley's law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24]. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was *mining* mode [@knight_etal21: 2]. The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.

As the variables in XRF analysis is given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods, and is managed by transforming the data using log-ratio transformation [@aitchison99]. There are currently three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted in the PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g. values below LOD, rounded zeros). Following log-transformation the data was then mean-centered and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].

\linebreak

## PCA {-}

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. Multivariate approaches, such as PCA,  were proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predictive models.

PCA analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].

\pagebreak

# Results {-}

\linebreak

## NIR - near infrared range {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d.filt <-
  points.d.tot %>% 
  filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                           "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56"))

#pass to vector so that length() can be used in text to quote total nr. of samples
spec.d.total <- 
  points.d.tot$sample_id

#pass to vector so that length() can be used in text to quote nr. of dark samples after SG analysis
spec.d.filtered <- 
  points.d.filt$sample_id
```

It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77]. Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra. This effect can be seen in @fig-nir-spectra-summary which shows a summary of the centered NIR spectra by color group. Apart from the dark material exhibiting a wider range of absorbance, likely owing some to the range of hues within the group, they generally show less distinct peaks than the material with a lighter hue. This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17: 78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21: 351-397; @schmid_etal22]. An example of this can be seen in @fig-nir-savitzky-transform, where the spectra from two different dark objects are compared. The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks. The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise. In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation. After reviewing the dark spectra, out of a total of `r length(spec.d.total)`, `r length(spec.d.filtered)` were kept in the dark group.

\linebreak

```{r}
#| label: fig-nir-spectra-summary
#| cache: true
#| fig-cap: "Summary of the raw spectra for each colour group. Note the differences in absorbance between the groups, with the dark group featuring a larger range of absorbance values as well as less distinct peaks."

quietly(source(here::here("analysis", "paper", "001-nir-spectra-summary.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-nir-savitzky-transform
#| cache: true
#| fig-cap: "Comparison of NIR spectra between two samples from the dark group. a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering."

quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.jpeg"), error = FALSE)
```

\linebreak

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (@fig-nir-savitzky-transform) & @fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77: 509; @clark_etal90: 12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current dataset the 2 200 nm band is typically the more intense band, although there are exceptions.

The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207]. Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2250 nm band for biotite [@duke94]. As can be seen in the average spectra in @fig-nir-peak-summary the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.

Two smaller features appear at around 1 065 nm and 1 300 nm (@fig-nir-savitzky-transform-alt), and are often vague and difficult to make out in the non-processed spectra. These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals. In studies of olivine the absorption region has been determined to be broad and centered around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13]. Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75: 104-106]. It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.

\linebreak

```{r}
#| label: fig-nir-savitzky-transform-alt
#| cache: true
#| fig-cap: "Savitzky-Golay transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups."

quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.jpeg"), error = FALSE)
```

```{r}
#| label: fig-nir-peak-summary
#| cache: true
#| fig-cap: "The main features in the NIR spectra can be related to overtones and vibration combinations of OH groups. The 1 400 nm and 1 900 nm peaks are indicative of molecular water, and common in quartz-based materials. The 2 200 nm and 2 350 nm bands indicate metal-OH combinations, where a more intense 2 200 nm band has typically been related to the presence of AlOH, and a more intense 2 350 nm band MgOH. Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)"

quietly(source(here::here("analysis", "paper", "002-nir-peak-summary.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| cache: true

quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
nir.pc1var <- nir.pc1var
nir.pc2var <- nir.pc2var
nir.pc3var <- nir.pc3var
```

\linebreak

The loading lines for the PCA can be seen in [Appendix A](#appendix-A). The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (@fig-nir-pca). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %). Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left. There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca:A.

While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) ([Appendix A](#appendix-a)). Adding PC 3 to a score plot (@fig-nir-pca:B), however, there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.

\linebreak

```{r}
#| label: fig-nir-pca
#| cache: true
#| fig-fullwidth: true
#| fig-height: 20
#| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separate three of the light samples."

knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)
```

\linebreak

## Raman {-}
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra-summary) & fig. \@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].

The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the expected position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

Aside from the above more intense features a couple of weaker peaks can also be seen in a number of samples. Some of these could potentially be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band. Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.

\linebreak

```{r}
#| label: fig-raman-spectra-summary
#| cache: true
#| fig-cap: "The non-processed Raman spectra visualised by colour group. Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^)."

quietly(source(here::here("analysis", "paper", "005-raman-spectra-summary.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-raman-peak-ident
#| cache: true
#| fig-cap: "The features in the Raman spectra can be related to quartz and carbon structures in the material. The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz. The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures."

quietly(source(here::here("analysis", "paper", "006-raman-peak-identification.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| cache: true

quietly(eval(parse(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8")))
raman.pc1var <- raman.pc1var
raman.pc2var <- raman.pc2var
raman.pc3var <- raman.pc3var
```

\linebreak

The loading lines for the PCA can be seen in [Appendix A](#appendix-A). The PCA model (@fig-raman-pca) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks. Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side. Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot. Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot. One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material. The second sample, however, stands out from the rest of the material. It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals. It does not fit with the "expected" image of a quartzite, and may be a case of mischaraterization. The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.

\linebreak

```{r raman-pca, fig.cap= "PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
#| label: fig-raman-pca
#| cache: true
#| fig-cap: "The features in the Raman spectra can be related to quartz and carbon structures in the material. The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz. The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures."

knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.jpg"), error = FALSE)
```

```{r}
quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
#| label: xrf-si
#| cache: true

#extract samples above 80% Si 
Si_90 <- Points.xrf %>% 
  filter(`Si` > 90)

#calculate percentage of samples above 80% Si
si_90 <- length(unique(Si_90$sample_id))
```

```{r}
#| label: xrf-pca-var
#| cache: true

quietly(source(here::here("analysis", "paper", "009-xrf-pca.R"), encoding = "UTF-8"))

xrf_var <- xrf.pc1var + xrf.pc2var

xrf_var_low <- xrf.pc1var.low + xrf.pc2var.low + xrf.pc3var.low + xrf.pc4var.low
```

## XRF {-}

The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18]. The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot). Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. While this could be due to the limitations of the instrumentation, it is perhaps more likely a result of the sampling. As quartzite is a metamorph rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%. ~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.

Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values. log10-transformation was used instead, producing a similar model, but that managed the loading for ~14~Si better. The PCA was run twice, firstly with the > 90% 14Si samples included and secondly excluded. The loadings can be seen in [Appendix A](#appendix-A). The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

As can be seen in the score plot (@fig-xrf-pca:A), the three K-means clusters remain separated. As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%). There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values. Loadings for elements that you would expect to follow changes in ~14~Si content also end up near cluster 1 (e.g. Al, Ca). Interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not showcase any typical quartz characteristics.

Cluster 2 is marked by a higher ~22~Ti, ~26~Fe and ~40~Zr content on average. It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. ~13~Al and ~19~K is similarly higher in cluster 2 than compared to cluster 3, which is marked by its, on average, lower elemental content than the other two clusters. As ~13~Al and ~19~K signifies a sedimentary depositional context that has been subjected to metamorphism, this may be a possible explanation.  That said, cluster 3 does features a more limited range of values between the samples. The material in cluster 3 also includes all of the *brecciated quartz* samples as well as other light - colourless objects.

In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%. Again, 3 clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot). Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (@fig-xrf-pca:B), than compared to the other clusters. The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples. Cluster 2, on the other hand, feature the main bulk of darker samples with higher ~22~Ti, ~26~Fe and ~40~Zr values. Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values. It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.

~38~Sr was detected at trace amounts. It is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B. Interestingly, nearly all samples with ~38~Sr are located in cluster 2 in model A.

\linebreak

```{r}
#| label: fig-xrf-violin-plot
#| cache: true
#| fig-cap: "Violin plot of the XRF values per element. Median and interquartile range represented with a boxplot, with the half-violin showing kernel density estimation. Values above LOD is noted by *n*. Of particular note is the bimodal distribution of Si values. Y and Zn were left out as they had less than 30 values above LOD."

quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "008-xrf-violin-plot.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-xrf-pca
#| cache: true
#| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster somewhat based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples also remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg."

knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.jpg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-xrf-kmeans-boxplot
#| cache: true
#| fig-cap: "Boxplot of XRF values (%) for model B, grouped by K-means cluster. Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with the same (e.g. Al, Ca). Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content."

knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.jpeg"), error = FALSE)
```

\linebreak

The K-means cluster analysis of model B is visualised in the NIRS model, which further emphasizes the variation within the general color groups (@fig-nir-xrf-kmeans). Cluster 2, with higher ~26~Fe and ~40~Zr values, mainly coincides with the darker material to the right side of the plot. This was expected as the trend along PC 1 in the NIR PCA mainly separates the material according to absorbance. Cluster 3 in turn coincides with the lighter material in the opposite side of the model. Cluster 1, however, is fairly evenly distributed within the model along PC 1 and PC 2.

The variation that occurs within the general colour groups can be further exemplified by looking closer at three samples selected from the dark group, that each belong to a different K-means cluster. They all share a similar hue with D/G peaks detected in the Raman analysis (@fig-nir-sample-comparison). All three samples have similar absorbance values, but sample 391 exhibits some differences in the spectra compared to the other two samples. Out of the three objects sample 391 seems to stand out more clearly from the others, with the XRF analysis being what mainly seems to separate the other two.

```{r}
#| label: fig-nir-xrf-kmeans
#| cache: true
#| fig-cap: "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points. Red points are samples that were excluded in the XRF PCA. The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot. This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD."

knitr::include_graphics(here::here("analysis", "figures", "009-nir-pca-xrf-kmeans.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-nir-sample-comparison
#| cache: true
#| tbl-cap: "Comparison of three dark samples from separate K-cluster groups (model B). Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples."

quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "010-nir-sample-comparison.jpeg"), error = FALSE)
```

\linebreak

# Discussion {-}

\linebreak
 \linebreak

The exploratory evaluation of spectroscopic data generated on quartz/quartzite points and preforms demonstrates that valuable information on the composition and structural properties of the material can be gained via non-destructive analysis. As the identified features are related to the diagenetic/paragenetic environment they should be useful in future survey efforts aimed at locating possible geological sources. This does not, however, address the issues related to the resource procurement strategies employed by people within the region during the 4 000 - 2 000 BP period, as the material may have been sourced from other glacial and postglacial landforms, into which the material may have been incorporated. One way of addressing this issue would be to construct a predictive model based solely on the spectroscopic fingerprints of the large unknown dataset, which could form the basis of spatial analysis and hypothesis testing.

One benefit of non-destructive screening is the relative ease with which large datasets can be generated. Although the resolution may be affected to some degree, it lends itself quite well to the exploratory approach and hypothesis generation. A non-destructive approach is quite effective in regions such as Northern Sweden, where little development has been made in regards to the chemical characterization of quartzite and there are many unknowns. There is now a better understanding of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as *quartzite*, at least within the Västerbotten area. Visually the material varies greatly, both in terms of texture as well as color. The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, but as was highlighted in the results, there seems to be some instances of possible mischaracterization due to this. It is an important reminder to be mindful when aggregating data where the analyst is not the primary producer.

As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former). The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus delimiting possible geological source areas. Usage of the term "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden. Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g. @gaweda_cebulak99; @ukar_cloos16]. Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.

Questions of metamorphic grade can also be raised based on the metal-OH features in the NIRS data. Should these be related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite). Not only does such information provide grounds for discussions on provenance, but it also aids in the construction of a predictive model. The main issue, however, seems to be the struggle with sampling dark material using NIRS (and to an extent Raman), which also feature many of the indicators for biotite in the material. Sampling in general is a core issue, as the question becomes how representative the results really are of the material/objects in question? This does not simply extend to the measurements, however, but also to the visual characterization. The division of a heterogeneous material into larger simplified color groups might give a mistaken impression of homogeneity. Should a light grey point with streaks of black mineral running through it be classified as a light or dark quartzite? The necessity for adopting such classifications in the past is understandable, a certain level of abstraction is necessary in higher level modelling after all [@sjolander22]. Spectroscopic screening, however, does provide a more quantitative and empirical approach to such tasks [@ramacciotti_etal22].

Due to the high representation of quartz and quartzite in the lithic assemblages dating to the Mesolithic - Bronze Age in Northern Sweden [@baudou78], it is vital that methods for characterization of these materials are developed. As they represent a primary lithic resource they hold a critical position in the study of resource procurement, trade and exchange and human mobility within the region (as well as many other regions [e.g. @prieto_etal19]). Previous studies have primarily relied on visual characterization following a system that was setup in order to manage the large quantities of data generated following the large scale development for hydroelectric power during the 1940s - 1980s [@biornstad06; @kaeck09: 53-55]. This was necessary in order to manage the overabundance of data and make sense of it at the time, and provided a lot of new insights [e.g. @forsberg85; @holm91; @lundberg97], but there has been little development in regards to questions of provenance of these materials since. This study addresses those needs and demonstrates the potential which exists for increasing the understanding of human resource management of locally available lithic material.

\linebreak

# Conclusion {-}

\linebreak
 \linebreak

The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification. Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials. The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material. Not only is there a need for methodological development in the study of quartzite in an archaeological context, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources. Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.

\linebreak

# CRediT authorship contribution statement {-}
Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft.

Johan Linderholm: Conceptualization, Methodology, Supervision

Paul Geladi: Formal analysis, Methodology

\linebreak

# Statements and Declarations

**Competing interests**
The authors declare no competing interests.

# Funding source {-}
This work was supported by PhD funding and facilities at the Environmental Archaeology Laboratory at Umeå University.

\linebreak

# Acknowledgements {-}

We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts. We would also like to acknowledge good cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.

\linebreak
 \linebreak

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

# Appendices{-}

\linebreak
 \linebreak

## Appendix A {#appendix-a}

Collection of loadings for each single instrumentation PCA.

\beginsupplement

\linebreak

```{r}
#| label: fig-nir-pca-loadings
#| fig-width: 20
#| fig-height: 20
#| cache: true
#| fig-cap: "PCA loadings for the first 3 principal components for the NIR model."

quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-raman-pca-loadings
#| fig-width: 20
#| fig-height: 20
#| cache: true
#| fig-cap: "PCA loadings for the first 3 principal components for the Raman model."

quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.jpeg"), error = FALSE)
```

\linebreak

```{r}
#| label: fig-xrf-pca-loadings-scree
#| fig-fullwidth: true
#| fig-fullheight: true
#| cache: true
#| fig-cap: "PCA Loadings and eigenvalues for the XRF data. a) eigenvalues and loadings for the full dataset (model A), b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded (model B)."

knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.jpeg"), error = FALSE)
```

\endsupplement

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2023-04-04 analysis/paper/001-artefact-size.R

@@ -1,0 +1,49 @@
+# Loading package
+library(ggplot2)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Filter data to focus on points and preforms made from quartz/quartzite material in Vilhelmina and Åsele parish
+Points <-
+  metadata.csv %>% 
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
+                  site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
[…]

### 2023-04-04 analysis/paper/001-nir-spectra-summary.R

@@ -1,0 +1,156 @@
+# Loading package
+library(data.table)
+library(ggplot2)
+library(gridBase)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
+
+#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
[…]

### 2023-04-04 analysis/paper/002-nir-dark-filtered-summary.R

@@ -1,0 +1,158 @@
+# Loading package
+library(gridBase)
+library(data.table)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
[…]

### 2023-04-04 analysis/paper/002-nir-peak-summary.R

@@ -1,0 +1,165 @@
+# Loading package
+suppressPackageStartupMessages(library(data.table))
+library(gridBase)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-04 analysis/paper/003-nir-savitzky-transform-alt.R

@@ -1,0 +1,83 @@
+#Load packages
+suppressPackageStartupMessages(library(data.table))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(reshape2))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
[…]

### 2023-04-04 analysis/paper/003-nir-savitzky-transform.R

@@ -1,0 +1,126 @@
+#Load packages
+suppressPackageStartupMessages(library(data.table))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(reshape2))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-04 analysis/paper/004-nir-pca-loadings.R

@@ -1,0 +1,104 @@
+#Load packages
+suppressPackageStartupMessages(library(plotly))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-04 analysis/paper/004-nir-pca.R

@@ -1,0 +1,144 @@
+#Load packages
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
+
+#Filter NIR data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-04 analysis/paper/005-raman-spectra-summary.R

@@ -1,0 +1,145 @@
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(spectrolab))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#merge raman de-trended data with metadata
+raman.baseline.merged <- 
+  as.data.frame(merge(metadata.csv, raman.averaged, by='sample_id'))
+
+Points.raman <-
+  raman.baseline.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-04 analysis/paper/006-raman-peak-identification.R

@@ -1,0 +1,158 @@
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(spectrolab))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+  select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
[…]

### 2023-04-04 analysis/paper/007-raman-pca-loadings.R

@@ -1,0 +1,158 @@
+#Load packages
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#transpose in prep for baseline de-trend
+raman.transposed <- 
+  t(raman.averaged)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
[…]

### 2023-04-04 analysis/paper/007-raman-pca.R

@@ -1,0 +1,152 @@
+#Load packages
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#transpose in prep for baseline de-trend
+raman.transposed <- 
+  t(raman.averaged)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"), col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"),
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
[…]

### 2023-04-04 analysis/paper/008-xrf-violin-plot.R

@@ -1,0 +1,164 @@
+library(ggplot2)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
[…]

### 2023-04-04 analysis/paper/009-xrf-pca.R

@@ -1,0 +1,676 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#### Load data ####
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#### Prepare dataset ####
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
[…]

### 2023-04-04 analysis/paper/010-nir-sample-comparison.R

@@ -1,0 +1,346 @@
+# Loading package
+library(png)
+library(loder)
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(cowplot))
+suppressPackageStartupMessages(library(data.table))
+suppressPackageStartupMessages(library(flextable))
+suppressPackageStartupMessages(library(grid))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
[…]

### 2023-04-04 README.Rmd
Updated readme

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023).
+(`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 

### 2023-04-07 analysis/paper/001-artefact-size.R
Updated scripts according to new metadata and data files

@@ -4,29 +4,19 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", dec = ".", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Filter data to focus on points and preforms made from quartz/quartzite material in Vilhelmina and Åsele parish
 Points <-
-  metadata.csv %>% 
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
-                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
-                  site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
[…]

### 2023-04-07 analysis/paper/001-nir-spectra-summary.R

@@ -6,94 +6,84 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", dec = ".", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
 
-#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+#Filter nir data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
-                  site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-07 analysis/paper/002-nir-dark-filtered-summary.R

@@ -5,81 +5,71 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", dec = ".", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
[…]

### 2023-04-07 analysis/paper/002-nir-peak-summary.R

@@ -5,91 +5,81 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-                  site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-                  site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
[…]

### 2023-04-07 analysis/paper/003-nir-savitzky-transform-alt.R

@@ -1,4 +1,5 @@
 #Load packages
+suppressPackageStartupMessages(library(ggtext))
 suppressPackageStartupMessages(library(data.table))
 suppressPackageStartupMessages(library(prospectr))
 suppressPackageStartupMessages(library(reshape2))
@@ -6,45 +7,35 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
-nir.averaged <- 
+nir.averaged <-
   aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                  site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                  site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
[…]

### 2023-04-07 analysis/paper/003-nir-savitzky-transform.R

@@ -1,4 +1,5 @@
 #Load packages
+suppressPackageStartupMessages(library(ggtext))
 suppressPackageStartupMessages(library(data.table))
 suppressPackageStartupMessages(library(prospectr))
 suppressPackageStartupMessages(library(reshape2))
@@ -6,58 +7,50 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
[…]

### 2023-04-07 analysis/paper/004-nir-pca-loadings.R

@@ -4,36 +4,28 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
-nir.averaged <- 
+nir.averaged <-
   aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2023-04-07 analysis/paper/004-nir-pca.R

@@ -4,34 +4,27 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.csv, by='sample_id'))
 
-#Filter NIR data to focus on points and preforms made from quartz/quartzite material 
+#Filter NIR data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2023-04-07 analysis/paper/005-raman-spectra-summary.R

@@ -1,82 +1,75 @@
-suppressPackageStartupMessages(library(ChemoSpec))
+#Load packages
 suppressPackageStartupMessages(library(spectrolab))
 suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(ggtext))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import raman data, set empty fields to NA
 raman.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+raman.averaged <-
+  aggregate(raman.csv[, 4:468], list(sample_id = raman.csv$sample_id), mean)
 
 #merge raman de-trended data with metadata
-raman.baseline.merged <- 
+raman.baseline.merged <-
   as.data.frame(merge(metadata.csv, raman.averaged, by='sample_id'))
 
 Points.raman <-
   raman.baseline.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
[…]

### 2023-04-07 analysis/paper/008-xrf-violin-plot.R

@@ -7,31 +7,23 @@ xrf.csv <-
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #merge XRF data with metadata
-xrf.merged <- 
+xrf.merged <-
   as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.xrf <-
   xrf.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
[…]

### 2023-04-07 analysis/paper/009-xrf-pca.R

@@ -15,39 +15,39 @@ nir.csv <-
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #merge XRF data with metadata
-xrf.merged <- 
+xrf.merged <-
   as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
 
 #### Prepare dataset ####
 
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material
 Points.xrf <-
   xrf.merged %>%
   filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
            site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
            site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
            site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2023-04-07 analysis/paper/010-nir-sample-comparison.R

@@ -7,10 +7,11 @@ suppressPackageStartupMessages(library(data.table))
 suppressPackageStartupMessages(library(flextable))
 suppressPackageStartupMessages(library(grid))
 suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(ggtext))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
@@ -25,180 +26,110 @@ xrf.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
-nir.averaged <- 
+nir.averaged <-
   aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
 
 #merge NIR data with metadata
-nir.merged <- 
+nir.merged <-
   as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
 
 #merge XRF data with metadata
-xrf.merged <- 
+xrf.merged <-
   as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+raman.averaged <-
+  aggregate(raman.csv[, 4:468], list(sample_id = raman.csv$sample_id), mean)
 
-#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+#Filter nir data to focus on points and preforms made from quartz/quartzite material
 Points.nir <-
   nir.merged %>%
[…]

### 2023-04-07 analysis/paper/006-raman-baseline.R
Added raman baseline script to make raman scripts more efficient

@@ -1,0 +1,70 @@
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <-
+  aggregate(raman.csv[, 4:468], list(sample_id = raman.csv$sample_id), mean)
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <-
+  raman.averaged %>%
+  select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <-
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <-
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
+
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and
+#return corrected spectra to spectra object
+raman.baseline <-
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#prevent baseline from plotting a figure
+dev.off()
+
+#spectra are stored in "Spectra" object, not a data frame
+#need to extract samples and frequencies from that object
+raman.tmp <-
+  cbind(raman.baseline$freq, t(raman.baseline$data))
[…]

### 2023-04-07 analysis/paper/006-raman-peak-identification.R

@@ -1,79 +1,17 @@
-suppressPackageStartupMessages(library(ChemoSpec))
 suppressPackageStartupMessages(library(spectrolab))
 suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(ggtext))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
-#Import raman data, set empty fields to NA
-raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
-
-#select sample_id and data columns and transpose in prep for baseline de-trend
-raman.transposed <- 
-  raman.averaged %>% 
-  select(sample_id, `92.88`:`2503.59`)
-
-raman.transposed <- 
-  t(raman.transposed)
-
-#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
-#ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
-
-#prepare vector with sample_id for reading raman data into spectra object
-sample_id <- 
-  raman.averaged$sample_id
-
-#read transposed raman data into spectra object
-raman <- suppressWarnings(matrix2SpectraObject(
-  gr.crit = sample_id,
-  gr.cols = c("auto"),
-  freq.unit ="Wavelength cm-1",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
-  sep = ";",
-  dec = ".",
-  chk = TRUE,
[…]

### 2023-04-07 analysis/paper/007-raman-pca-loadings.R

@@ -1,76 +1,19 @@
 #Load packages
-suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(ggtext))
 suppressPackageStartupMessages(library(factoextra))
 suppressPackageStartupMessages(library(prospectr))
 suppressPackageStartupMessages(library(tidyverse))
 
-#Import raman data, set empty fields to NA
-raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
+#Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
-
-#transpose in prep for baseline de-trend
-raman.transposed <- 
-  t(raman.averaged)
-
-#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
-#ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
-
-#prepare vector with sample_id for reading raman data into spectra object
-sample_id <- 
-  raman.averaged$sample_id
-
-#read transposed raman data into spectra object
-raman <- suppressWarnings(matrix2SpectraObject(
-  gr.crit = sample_id,
-  gr.cols = c("auto"),
-  freq.unit ="Wavelength cm-1",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
-  sep = ";",
-  dec = ".",
-  chk = TRUE,
[…]

### 2023-04-07 analysis/paper/007-raman-pca.R

@@ -1,76 +1,18 @@
 #Load packages
-suppressPackageStartupMessages(library(ChemoSpec))
 suppressPackageStartupMessages(library(factoextra))
 suppressPackageStartupMessages(library(prospectr))
 suppressPackageStartupMessages(library(tidyverse))
 
-#Import raman data, set empty fields to NA
-raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
+#Import descriptive metadata
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), colClasses=c("sample_id"="character"), dec = ".", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
-
-#transpose in prep for baseline de-trend
-raman.transposed <- 
-  t(raman.averaged)
-
-#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
-#ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"), col.names = FALSE, sep = ";", dec = )
-
-#prepare vector with sample_id for reading raman data into spectra object
-sample_id <- 
-  raman.averaged$sample_id
-
-#read transposed raman data into spectra object
-raman <- suppressWarnings(matrix2SpectraObject(
-  gr.crit = sample_id,
-  gr.cols = c("auto"),
-  freq.unit ="Wavelength cm-1",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"),
-  sep = ";",
-  dec = ".",
-  chk = TRUE,
-  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
-
-#call on relevant baseline method (modified polynomial fitting) from baseline package and 
[…]

### 2023-04-07 analysis/paper/paper.qmd
Updated manuscript, mainly figure sizing and references

@@
 date-format: long
+crossref: 
+  fig-title: '**Fig.**'
+  fig-labels: '**arabic**'
+  fig-prefix: Fig.
 format:
@@
   fig-path: "../figures/"
-  fig-dpi: 600
+  fig-dpi: 300
 filters:
@@
 metadata.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", dec = ".", colClasses=c("sample_id"="character"), header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
@@
   metadata.csv %>% 
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-                site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-                site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-                site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-                site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
-                site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-                site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
-                site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
-                site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
-                site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
+  dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
@@
 Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."
-#| fig-height: 15
-#| fig-fullwidth: true
 
@@
 
-(ref:artefact)
-
 \linebreak
@@
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.618
+#| fig-align: center
 #| fig-cap: "Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts.
@@
 Note that the term *point* does not differentiate between spearheads or arrowheads."
+quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
+#knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)
 
-quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)
+fig
 ```
@@
   metadata.csv %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
-           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
-           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
-           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
-           site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
+  filter( type == "Point" | type == "Point fragment" | type == "Preform", 
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
@@
 
-```{r}
+```{r fig.width=12, fig.height = 5}
 #| label: fig-nir-spectra-summary
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "Summary of the raw spectra for each colour group.
@@
 quietly(source(here::here("analysis", "paper", "001-nir-spectra-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.jpeg"), error = FALSE)
+
+fig
 ```
@@
 
-```{r}
+```{r fig.width=12, fig.height = 5}
 #| label: fig-nir-savitzky-transform
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "Comparison of NIR spectra between two samples from the dark group.
@@
 quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.jpeg"), error = FALSE)
+
+fig
 ```
@@
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.6
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "Savitzky-Golay transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups."
@@
 quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.jpeg"), error = FALSE)
+
+fig
 ```
@@
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "The main features in the NIR spectra can be related to overtones and vibration combinations of OH groups.
@@
 quietly(source(here::here("analysis", "paper", "002-nir-peak-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.jpeg"), error = FALSE)
+
+fig
 ```
@@
 ```{r}
+#| label: pca-var
 #| cache: true
@@
 Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left.
-There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca:A.
+There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca :A.
 
 While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) ([Appendix A](#appendix-a)).
-Adding PC 3 to a score plot (@fig-nir-pca:B), however, there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.
+Adding PC 3 to a score plot (@fig-nir-pca :B), however, there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.
 
@@
 #| cache: true
-#| fig-fullwidth: true
-#| fig-height: 20
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separate three of the light samples."
+#| 
+quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
+#knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)
 
-knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)
+fig
 ```
@@
 ## Raman {-}
-There are two distinguishing features present in the Raman data (fig.
-\@ref(fig:raman-spectra-summary) & fig.
-\@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
+There are two distinguishing features present in the Raman data (@fig-raman-spectra-summary) & @fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
 The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
@@
 #| cache: true
-#| fig-cap: "The non-processed Raman spectra visualised by colour group. Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^)."
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
+#| fig-align: center
+#| fig-cap: "The non-processed Raman spectra visualised by colour group. Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^)"
 
 quietly(source(here::here("analysis", "paper", "005-raman-spectra-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.jpeg"), error = FALSE)
+
+fig
 ```
@@
 ```{r}
+#| label: raman-baseline
+#| cache: true
+
+#baseline correct raman data
+quietly(source(here::here("analysis", "paper", "006-raman-baseline.R"), encoding = "UTF-8"))
+```
+
+```{r}
 #| label: fig-raman-peak-ident
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
+#| fig-align: center
 #| fig-cap: "The features in the Raman spectra can be related to quartz and carbon structures in the material.
 The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
-The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures."
+The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures"
 
+#generate figure
 quietly(source(here::here("analysis", "paper", "006-raman-peak-identification.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.jpeg"), error = FALSE)
+
+fig
 ```
@@
 ```{r}
+#| label: raman-var
 #| cache: true
@@
 ```
-
-\linebreak
 
@@
 It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals.
-It does not fit with the "expected" image of a quartzite, and may be a case of mischaraterization.
+It does not fit with the "expected" image of a quartzite and may be a case of erroneous classification.
 The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.
@@
 
-```{r raman-pca, fig.cap= "PCA score plot for Raman data.
+```{r}
+#| label: fig-raman-pca
+#| cache: false
+#| cache-lazy: false
+#| fig-cap: "PCA score plot for Raman data.
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
-#| label: fig-raman-pca
-#| cache: true
-#| fig-cap: "The features in the Raman spectra can be related to quartz and carbon structures in the material.
-The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
-The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures."
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster"
 
+#quietly(source(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8"))
 knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.jpg"), error = FALSE)
 ```
+
+\linebreak
+
+## XRF {-}
 
@@
 
-## XRF {-}
-
 The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
@@
 
+\linebreak
+
+```{r}
+#| label: fig-xrf-violin-plot
+#| cache: true
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "70%"
+#| fig-align: center
+#| fig-cap: "Violin plot of the XRF values per element.
+Median and interquartile range represented with a boxplot, with the half-violin showing kernel density estimation.
+Values above LOD is noted by *n*.
+Of particular note is the bimodal distribution of Si values.
+Y and Zn were left out as they had less than 30 values above LOD."
+
+quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
+
+fig
+```
+
+\linebreak
+
 Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values.
@@
 
-As can be seen in the score plot (@fig-xrf-pca:A), the three K-means clusters remain separated.
+As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated.
 As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
@@
 Again, 3 clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot).
-Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (@fig-xrf-pca:B), than compared to the other clusters.
+Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (@fig-xrf-pca :B), than compared to the other clusters.
 The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples.
@@
 Interestingly, nearly all samples with ~38~Sr are located in cluster 2 in model A.
-
-\linebreak
-
-```{r}
-#| label: fig-xrf-violin-plot
-#| cache: true
-#| fig-cap: "Violin plot of the XRF values per element.
-Median and interquartile range represented with a boxplot, with the half-violin showing kernel density estimation.
-Values above LOD is noted by *n*.
-Of particular note is the bimodal distribution of Si values.
-Y and Zn were left out as they had less than 30 values above LOD."
-
-quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "008-xrf-violin-plot.jpeg"), error = FALSE)
-```
-
-\linebreak
 
@@
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "100%"
 #| fig-cap: "Boxplot of XRF values (%) for model B, grouped by K-means cluster.
@@
 
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.jpeg"), error = FALSE)
+#Since knitr commands for adjusting knitr imported figures don't work in quarto, and the xrf script generates multiple figures, the easiest way to manage this is saving the boxplot
+#as an object to derived_data when running xrf-pca-var above and then loading it here and outputting it in manuscript
+fig <- 
+  readRDS(file = here::here("analysis", "data", "derived_data", "009-kmeans-box-modelb.rds"))
+
+fig
 ```
@@
 #| cache: true
+#| fig-width: 10
+#| fig-asp: 0.687
+#| out-width: "70%"
 #| fig-cap: "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points.
@@
 
-knitr::include_graphics(here::here("analysis", "figures", "009-nir-pca-xrf-kmeans.jpeg"), error = FALSE)
+#Since knitr commands for adjusting knitr imported figures don't work in quarto, and the xrf script generates multiple figures, the easiest way to manage this is saving the boxplot
+#as an object to derived_data when running xrf-pca-var above and then loading it here and outputting it in manuscript
+fig <- 
+  readRDS(file = here::here("analysis", "data", "derived_data", "009-kmeans-nir-pca.rds"))
+
+fig
 ```
@@
 ```{r}
-#| label: fig-nir-sample-comparison
 #| cache: true
-#| tbl-cap: "Comparison of three dark samples from separate K-cluster groups (model B).
-Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples."
 
 quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "010-nir-sample-comparison.jpeg"), error = FALSE)
 ```
+
+![Comparison of three dark samples from separate K-cluster groups (model B).
+Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples](../figures/010-nir-sample-comparison.jpeg){#fig-nir-sample-comparison width=7.87in, height=7.87in}
 
@@
 Visually the material varies greatly, both in terms of texture as well as color.
-The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, but as was highlighted in the results, there seems to be some instances of possible mischaracterization due to this.
+The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, but as was highlighted in the results, some may have been wrongly classified due to this.
 It is an important reminder to be mindful when aggregating data where the analyst is not the primary producer.
@@
 
-# CRediT authorship contribution statement {-}
+# Author Contributions {-}
 Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft.
@@
 **Competing interests**
+
 The authors declare no competing interests.
+
+\linebreak
 
@@
 We would also like to acknowledge good cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.
-
-\linebreak
- \linebreak
 
@@
 
-# Appendices{-}
-
-\linebreak
- \linebreak
-
-## Appendix A {#appendix-a}
+# Appendix A {#appendix-a}
 
@@
 #| label: fig-nir-pca-loadings
-#| fig-width: 20
-#| fig-height: 20
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
 #| fig-cap: "PCA loadings for the first 3 principal components for the NIR model."
@@
 quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.jpeg"), error = FALSE)
+
+fig
 ```
@@
 #| label: fig-raman-pca-loadings
-#| fig-width: 20
-#| fig-height: 20
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
 #| fig-cap: "PCA loadings for the first 3 principal components for the Raman model."
@@
 quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.jpeg"), error = FALSE)
+#knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.jpeg"), error = FALSE)
+
+fig
 ```
@@
 #| label: fig-xrf-pca-loadings-scree
-#| fig-fullwidth: true
-#| fig-fullheight: true
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 0.8
+#| out-width: "70%"
 #| fig-cap: "PCA Loadings and eigenvalues for the XRF data.
@@
 
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.jpeg"), error = FALSE)
+fig <- 
+  readRDS(file = here::here("analysis", "data", "derived_data", "009-xrf-loadings.rds"))
+
+fig
 ```
@@
 \newpage
+
+#Data availability
+
+The data and code used in the analysis is available at DOI XXX.
 

### 2023-04-26 analysis/paper/paper.qmd
new csl template

@@
 While this could be due to the limitations of the instrumentation, it is perhaps more likely a result of the sampling.
-As quartzite is a metamorph rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement.
+As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement.
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.

### 2023-04-27 analysis/paper/009-xrf-pca.R
slight layout update

@@ -643,10 +643,10 @@ fig4 <-
   ggnewscale::new_scale_color() +
   geom_point(data=subset(nir, is.na(variables)),
              aes(x=x, y=y, col="#D62728FF")) +
-  scale_color_manual(name="Missing XRF data",
+  scale_color_manual(name="Objects missing XRF data",
                      labels="NA",
                      values="#D62728FF") +
-  guides(color = guide_legend(override.aes = list(shape = 22,
+  guides(color = guide_legend(override.aes = list(shape = 21,
                                                  fill = "#D62728FF",
                                                  ncol = 1,
                                                  color = "black",

### 2023-05-12 analysis/paper/paper.qmd
Updated manuscript and restructured

@@
       institute: [SLU]
+      correspondence: false
+  - Philip I. Buckland:
+      email: philip.buckland@umu.se
+      institute: [UmU]
       correspondence: false
@@
 bibliography: references.bib
-csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
+csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
 abstract: |
@@
 Provenance of locally available geological material thus becomes complex and fraught with uncertainty.
-Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region.
-However, there is a severe lack of methodological development in the characterization of the same these materials.
-125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden.
-The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
-Evaluation of the spectra and exploratory analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin and potentially the metamorphic grade.
-These features indicate the presence of graphite, muscovite, and biotite.
-Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region.
-There also seems to be a good ground for the construction of a predictive model that could classify the quartzite based on its chemical characteristic.
-Such a model could prove useful in spatial analysis and testing of models of raw material management.
+Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region.
+However, there is a severe lack of methodological development in the characterization of these materials.
+125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden.
+The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
+Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin.
+The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade.
+In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region.
+They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic.
+Such a model would prove useful in future spatial analysis and testing of models of raw material management.
 keywords: |
   Lithics; Provenance; Quartzite; Spectroscopy; Sweden
-highlights: |
-  These are the highlights. 
 ---
@@
 
-Highlights: `r rmarkdown::metadata$highlights`
+Highlights:
+
+* Non-destructive analysis of prehistoric bifacial points made in quartzite using near-infrared spectroscopy, Raman scattering and X-Ray Fluourescence. 
+* Each method detects differences in the structural and chemical characteristics in the material, but on their own the resolution is too low for predictive modelling. 
+* Data fusion techniques are recommended for predictive modelling using non-destructive spectroscopy on quartzite material.
 
@@
 
-Petrographic characterization and provenance studies are instrumental for inferences of trade and exchange among prehistoric societies [@tykot03].
-The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
-The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources.
+Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03].
+The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery.
+The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management.
 Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17].
-The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
-The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent has been highlighted by Prieto et al.
+This focus has led to a bias, however, which has been highlighted by Prieto et al.
 [-@prieto_etal19: 15].
-In Northern Sweden both quartz and quartzite dominate the lithic assemblages and have been used in studies of settlements, exchange and mobility [@baudou78].
-While quartz and quartzite technology have been the target of several studies over the years, there has been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-M While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden haves been put forth [@halen94: 115-122; @lundberg97: 161-171].
-On the other hand there haves been little development on the topic of quartz and quartzite [@holm91: 23-24].
+Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+
+In Northern Sweden, quartz and quartzite dominate the lithic assemblages and have been used in settlement, exchange and mobility studies [@baudou78].
+Models have been proposed for raw material sourcing and trade and exchange for the less common slate and flint finds within Northern Sweden [@halen94: 115-122; @lundberg97: 161-171].
+There has, however, been little development in the petrographic characterization and provenance study of quartz and quartzite [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+).
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The abundance of moraine deposits and the degree to which geological materials have been mixed and transported necessitates considerable effort to reconstruct the events that have formed the current landscape.
-Existing theories on lithic material sourcing within the North Swedish region tend to describe strategies regarding direct procurement from the mountain zone [@forsberg85; -@forsberg12].
-This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06: 249-250].
-Due to this ambiguous nature of availability of raw materials, provenance studies are complex.
+Northern Sweden is extensively covered by glacial till, and whilst there are identifiable source areas (e.g.
+small bedrock quarries), there is an extremely large number of other potential sources of raw materials to be found in moraine deposits.
+The few existing theories on lithic material sourcing within the North Swedish region describe strategies for direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+This does not, however, account for the possibility that the material was sourced locally around the site, as part of an embedded adaptation strategy [@binford79; @rankama_etal06: 249-250].
+A new approach to provenance studies is therefore required in this region.
 
 Prieto et al.
-[-@prieto_etal19: 15; -@prieto_etal20: 32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature.
-Because of this , and different researchers have thus emphasized different different features as being characteristics of the material [@howard05].
-Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05: 708] also emphasises the need for field-based methodologies.
-Spectroscopic surface measurements are fast, and can be adapted for use in the field.
-In the case of archaeological research there are situations where destructive techniques, such as thin-section analysis, are not suitable or applicable.
-Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
+[-@prieto_etal19: 15; -@prieto_etal20: 32] note the ambiguous usage of the term quartzite in archaeology, with it often being used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+When defining quartzite, different researchers have emphasised different features as being characteristics of the material [@howard05].
+Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures.
+The need for field-based methodologies is, however, stressed by Howard [-@howard05: 708].
+Spectroscopic surface measurements are fast and can be adapted for use in the field.
+However, artefact collections (museum, private, etc.) also represent a valuable resource and asset, with easy access to large potential datasets with a wide geographical distribution.
+Destructive analysis on material from these collections, which are often unique and include sensitive cultural heritage material, are not always suitable or permitted, and thus there is a clear need for the development and testing of non-destructive techniques such as those presented in this study.
 
-X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data.
-The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
+In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden.
+X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data.
+The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
 Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
@@
 
-In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
-The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (@fig-distribution-map).
-The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources.
-The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management.
-Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models.
-The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.
+This paper presents an assessment of the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources.
+The geological complexity of both the region and the material necessitates alternative approaches to studies of raw material management.
+Without relying on well-delimited source areas for geological material, a predictive model capable of separating different “classes” of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models.
+The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden between 4 000 and 2 000 BP.
 
@@
 
-The current study includes spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map).
-The sites are distributed along the Ångerman river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south.
-The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
-No distinction is made between arrow- and spearheads.
+This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite.
+The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map).
+The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south.
+The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.
 
-A number of these points have, according to prior classifications, been worked using bifacial technology.
-The geographically closest chronology for bifacial points can be found in southeast Norway [@mjaerum12], but so far few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89], ).
-These attempts suggests a period of use between ca.
+The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89].
+The latter suggest a period of use between ca.
 3 800 - 2 000 BP [@kaeck09: 60].
-Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].
+Studies of raw material use at sites dating to this period seem to indicate an increase in the use of quartzite when compared to earlier periods [@baudou78].
 
@@
 #| cache-lazy: false
-#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine.
+#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
+The glacial disturbance is visible in the widely distributed moraine.
 Tectonic symbology after [@gee_stephens20].
-Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."
+Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"
 
@@
 
-The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe.
+The dataset includes three main groups of petrographic material, as classified by archaeologists: quartz, quartzite, and brecciated quartz (Swe.
 "*brecciekvarts*").
-Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection.
+Artefacts that had no prior classification were sorted into either the quartz or quartzite category based on visual inspection.
 This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples.
-The inclusion of quartz and brecciated quartz in the study serves two purposes.
-The first being to test the potential for separation of quartz from quartzite based on non-destructive analysis of their chemical and structural characteristics, and the second related purpose is to account for potential mistakes in the original classification.
+The inclusion of quartz and brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
 
-When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
-Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis.
-*Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
+In his analysis of the petrographic material in the Early Norrland project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
+Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis.
+*Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
 @holm91: 24].
-As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.
+As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.
 
@@
 #| cache-lazy: false
-#| fig-cap: "The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
+#| fig-cap: "Examples of the bifacial points studied.
+The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
 Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
-Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 2-76 mm.
-"
+Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"
 
@@
 
-All measurements were performed at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/).
-The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens.
-The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape (@fig-artefacts - @fig-artefact-size) (see Supplementary Materials).
-The study is non-destructive sampling was performed on the surface of a material that is not uniform.
-The material ranges from fairly small and flat point fragments, to large and rough preforms.
-Therefore, the signal integration could at times be difficult, which in turn could lead to scattering effects and saturation.
+All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/).
+The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens.
+The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data).
+The artefacts range from small and flat point fragments, to large and rough preforms.
+Therefore, the signal integration could at times be difficult, leading to scattering effects and saturation.
+For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult.
+When sampling the probe was in contact with the surface of the artefact at as flat a surface as possible.
+If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g.
+the dark mineral veins in @fig-artefact-size).
 
-Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
-Sampling materials that featured high heterogeneity the sample point was focused on the surrounding matrix, away from deviating inclusions.
-An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
-This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind.
-The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.
+For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object.
+The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution.
+The risk of misinterpretation is, however, somewhat compensated by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source.
+The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.
 
-The artefacts features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
-While the colour of the material has been classified according to the Munsell system, for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
+In colour, the artefacts exhibit a mainly black or grey hue, with a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
+While the colour of the material has been classified according to the Munsell system (see supplementary data), for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
 In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.
@@
 #| fig-align: center
-#| fig-cap: "Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts.
-8 objects have no recorded thickness, which ranges between 3-76 mm.
-Note that the term *point* does not differentiate between spearheads or arrowheads."
+#| fig-cap: "Scatter plot of length and width measurements (mm) for the sampled artefacts.
+8 artefacts have no recorded thickness, which ranges between 3-76 mm.
+Note that the term point does not differentiate between spear points or arrowheads"
 quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
@@
 
-### Near Infrared {-}
-NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
-Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near infrared region.
-The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify the bond.
-Overtones occur at frequencies that are a multiple of the infrared vibration.
-Most of them end up in the near infrared region [@osborne_etal93].
+### Near Infrared Spectroscopy (NIRS) {-}
 
+Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near-infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
+The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify it.
 Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
 The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].
+Overtones occur at frequencies that are a multiple of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].
 
-NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe.
+NIR analysis was conducted with an Analytical Spectral Device (ASD) LabSpec 4, using a contact probe (spot size: 10 mm).
 The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp.
@@
 Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
-The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before PCA modelling.
+The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before Principal Component Analysis (PCA) modelling.
 
@@
 ### Raman {-}
-Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01].
-However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect.
+Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared region [@nafie01].
+Raman spectroscopy is based on the *Raman scattering* effect.
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
@@
 
-### X-Ray Fluourescence {-}
-XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
-Spectral features in XRF spectrometry can be related to certain elements according to Moseley's law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24].
+### X-Ray Fluorescence (XRF) {-}
+XRF spectroscopy is a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
+Spectral features in XRF spectrometry can be related to certain elements according to Moseley’s law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24].
 Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159].
-In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample.
+In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample.
 One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.
+XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand.
+The reference calibration used for element quantification was mining mode [@knight_etal21: 2].
+The integration time was set to 130 seconds.
 
-XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
-The reference calibration used for element quantification was *mining* mode [@knight_etal21: 2].
-The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light).
-Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.
+As the variables in XRF analysis are given as proportions of a constant sum (e.g.
+percentage, parts per million (ppm)), this creates something referred to as the closed sum effect [@pawlowsky-glahn_egozcue06].
+This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20].
+This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99].
+Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15].
+In the PCA, the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
+@reimann_etal12].
+With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect (REF).
 
-As the variables in XRF analysis is given as proportions of a constant sum (e.g.
-percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06].
-This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20].
-This has implications for classical statistical methods, and is managed by transforming the data using log-ratio transformation [@aitchison99].
-There are currently three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15].
-In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
-@reimann_etal12].
-This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.
-
-As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA.
+As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA.
 A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement.
-Since missing values are not permitted in the PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15].
-*lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g.
+Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15].
+*lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g.
 values below LOD, rounded zeros).
-Following log-transformation the data was then mean-centered and scaled to unit variance before PCA.
+Following log-transformation the data was mean-centred and scaled to unit variance before PCA.
 Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].
@@
 
-## PCA {-}
+## Principal Component Analysis (PCA) {-}
 
-The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20].
-Multivariate approaches, such as PCA,  were proven useful for exploring complex spectroscopic datasets [@sciuto18].
+The data were evaluated using PCA, an unsupervised exploratory technique for reducing the dimensionality of a dataset  [@geladi_linderholm20].
+Multivariate approaches, such as PCA, have been proven useful for exploring complex spectroscopic datasets [@sciuto18].
 Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material.
-These groupings can then be followed up with supervised methods, and aid in the construction of predictive models.
+These groupings can then be followed up with supervised methods, and aid in the construction of predictive models for classifying unknown materials according to previously defined groups.
 
@@
 
+Spectra and PCA loadings for each instrumentation can be found in the supplementary data. An open repository with the code used in the analysis is linked in the *Data availability* section.
+
 \linebreak
 
-## NIR - near infrared range {-}
+## Near Infrared Spectroscopy (NIRS) {-}
 ```{r nir-sample-count, results='hide'}
@@
 
-It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
-Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra.
-This effect can be seen in @fig-nir-spectra-summary which shows a summary of the centered NIR spectra by color group.
-Apart from the dark material exhibiting a wider range of absorbance, likely owing some to the range of hues within the group, they generally show less distinct peaks than the material with a lighter hue.
-This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.
+NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
+Similarly, it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra.
+The dark material exhibits a wider range of absorbance, most likely due to the range of hues within the group, and generally shows less distinct peaks than the material with a lighter hue.
 
-The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17: 78-81].
+The spectra were evaluated using Savitzky-Golay (SG) filtering, a smoothing technique widely applied in spectroscopy [@chryssikos_gates17: 78-81].
 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21: 351-397; @schmid_etal22].
-An example of this can be seen in @fig-nir-savitzky-transform, where the spectra from two different dark objects are compared.
-The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks.
-The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise.
-In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation.
-After reviewing the dark spectra, out of a total of `r length(spec.d.total)`, `r length(spec.d.filtered)` were kept in the dark group.
+After reviewing the *dark* group using SG filtering, `r length(spec.d.filtered)` out of a total of `r length(spec.d.total)` artefacts were kept in the analysis.
+
+Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm.
+These features are all the result of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19].
+OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region.
+While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH [@hunt77: 509; @clark_etal90: 12664].
+In the current dataset the 2 200 nm band is typically the more intense band.
+
+The 2 200 nm and 2 350 nm absorption bands can be found in mica, such as muscovite and biotite, which can be found in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207].
+Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2 250 nm band for biotite [@duke94].
+As can be seen in the average spectra in @fig-nir-spectra-summary, the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.
+
+Two smaller features were recorded, albeit vague and difficult to make out in the non-processed spectra, at 1 065 nm and 1 300 nm (@fig-nir-spectra-summary).
+These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
+In studies of olivine the absorption region has been determined to be broad and centred around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13].
+Alternatively, weak water features have also been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water.
+However, these overlap with the Fe^2+^ bands in, for example, biotite [@adams75: 104-106].
+It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.
 
 \linebreak
-
+<!--
 ```{r fig.width=12, fig.height = 5}
@@
 
-Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm.
-These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (@fig-nir-savitzky-transform) & @fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19].
-OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region.
-As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77: 509; @clark_etal90: 12664].
-While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH.
-In the current dataset the 2 200 nm band is typically the more intense band, although there are exceptions.
-
-The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207].
-Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2250 nm band for biotite [@duke94].
-As can be seen in the average spectra in @fig-nir-peak-summary the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.
-
-Two smaller features appear at around 1 065 nm and 1 300 nm (@fig-nir-savitzky-transform-alt), and are often vague and difficult to make out in the non-processed spectra.
-These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
-In studies of olivine the absorption region has been determined to be broad and centered around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13].
-Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75: 104-106].
-It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.
-
-\linebreak
-
 ```{r}
@@
 ```
+-->
 
@@
 
-The loading lines for the PCA can be seen in [Appendix A](#appendix-A).
-The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (@fig-nir-pca).
+The loading lines for the PCA can be seen in the supplementary data.
+The first principal component (PC 1) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (@fig-nir-pca).
 This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative.
 A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %).
-Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left.
+Objects with a white or light grey hue gravitate towards the bottom left of the score plot, while material that displays some translucence cluster towards the top left.
 There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca :A.
 
-While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) ([Appendix A](#appendix-a)).
-Adding PC 3 to a score plot (@fig-nir-pca :B), however, there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.
+Whilst the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1).
+Adding PC 3 to a score plot (@fig-nir-pca :B), however, provides little new information apart from a couple of samples, mainly from the light group, that break away from the larger cluster.
 
@@
 #| fig-align: center
-#| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separate three of the light samples."
+#| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples"
 #| 
@@
 ## Raman {-}
-There are two distinguishing features present in the Raman data (@fig-raman-spectra-summary) & @fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
-The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
+There are two distinguishing features present in the Raman data (@fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
+The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, , are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
 
-The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples.
-These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99].
-The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
-Whereas the 2^nd^ peak falls within the expected position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
-It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280].
+The second feature shows as two peaks at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominates the spectra in terms of intensity for the darker samples.
+These positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99].
+The D band occurs as a result of disordered carbon (C) structures, whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
+The G band peak falls within the expected position, whereas the position of the D band is more commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
+It has, however, been documented that the dispersive behaviour of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280].
 This could thus explain the discrepancy between the observed peak in the dataset and the literature.
 
-Aside from the above more intense features a couple of weaker peaks can also be seen in a number of samples.
-Some of these could potentially be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band.
-Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
+In addition to the more intense features described above, a number of weaker peaks can also be seen in the spectra.
+Some of these could be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band.
+that is obscured by the D band.
+Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been identified as one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
 
 \linebreak
-
+<!--
 ```{r}
@@
 ```
-
+-->
  \linebreak
@@
 #| fig-align: center
-#| fig-cap: "The features in the Raman spectra can be related to quartz and carbon structures in the material.
-The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
+#| fig-cap: "TAveraged and baseline corrected Raman spectra for the light and dark groups.
+The features can be related to quartz and carbon structures in the material.
+The initial three peaks in 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
 The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures"
@@
 
-The loading lines for the PCA can be seen in [Appendix A](#appendix-A).
 The PCA model (@fig-raman-pca) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks.
@@
 It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals.
-It does not fit with the "expected" image of a quartzite and may be a case of erroneous classification.
+It does not fit with the “expected” image of a quartzite and may be a case of erroneous classification.
 The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.
@@
 
-## XRF {-}
+## X-Ray Fluorescence (XRF) {-}
 
@@
 
-The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
+The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
 The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18].
 The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot).
-Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
-While this could be due to the limitations of the instrumentation, it is perhaps more likely a result of the sampling.
+The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
+Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling.
 As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement.
@@
 ~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
+
+Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values.
+Instead log10-transformation was used, producing a similar model, but one that managed the loading for ~14~Si better.
+The PCA was run twice, firstly with the > 90% ~14~Si samples included and secondly excluded.
+The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
+K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition.
+Using the elbow graph method 3 clusters were deemed suitable for the analysis.
 
@@
 #| fig-align: center
-#| fig-cap: "Violin plot of the XRF values per element.
-Median and interquartile range represented with a boxplot, with the half-violin showing kernel density estimation.
-Values above LOD is noted by *n*.
+#| fig-cap: "Violin plots of the XRF values per element.
+Median and interquartile range is represented with a boxplot, with the half-violin showing kernel density estimation.
+Number of values above LOD is noted by n.
 Of particular note is the bimodal distribution of Si values.
-Y and Zn were left out as they had less than 30 values above LOD."
+Y and Zn were left out as they had less than 30 values above LOD"
 
@@
 
-Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values.
-log10-transformation was used instead, producing a similar model, but that managed the loading for ~14~Si better.
-The PCA was run twice, firstly with the > 90% 14Si samples included and secondly excluded.
-The loadings can be seen in [Appendix A](#appendix-A).
-The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
-K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition.
-Using the elbow graph method 3 clusters were deemed suitable for the analysis.
+As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated.
+Cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
+There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values.
+Loadings for elements expected to follow changes in 14Si content also end up near cluster 1 (e.g.
+Al, Ca).
+One of the “quartz” labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not display any typical quartz characteristics.
 
-As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated.
-As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
-There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values.
-Loadings for elements that you would expect to follow changes in ~14~Si content also end up near cluster 1 (e.g.
-Al, Ca).
-Interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not showcase any typical quartz characteristics.
+Cluster 2 is distinguished by a higher ~22~Ti, ~26~Fe and ~40~Zr content.
+It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue.
+~13~Al and ~19~K is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
+As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism (REF FROM JOHAN), and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
+Cluster 3 features a more limited range of values between the samples and includes all the brecciated quartz.
 
-Cluster 2 is marked by a higher ~22~Ti, ~26~Fe and ~40~Zr content on average.
-It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
-~13~Al and ~19~K is similarly higher in cluster 2 than compared to cluster 3, which is marked by its, on average, lower elemental content than the other two clusters.
-As ~13~Al and ~19~K signifies a sedimentary depositional context that has been subjected to metamorphism, this may be a possible explanation.
-That said, cluster 3 does features a more limited range of values between the samples.
-The material in cluster 3 also includes all of the *brecciated quartz* samples as well as other light - colourless objects.
-
-In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
-Again, 3 clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot).
-Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (@fig-xrf-pca :B), than compared to the other clusters.
+In the second PCA model (model B) four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
+Three clusters were deemed suitable for the K-means analysis (@fig-xrf-kmeans-boxplot).
+The range of values is more limited in cluster 3, with overall lower values than compared to the other clusters (@fig-xrf-pca :B).
 The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples.
-Cluster 2, on the other hand, feature the main bulk of darker samples with higher ~22~Ti, ~26~Fe and ~40~Zr values.
+Cluster 2 features the main bulk of darker samples with higher  ~22~Ti, ~26~Fe and ~40~Zr values.
 Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values.
@@
 ~38~Sr was detected at trace amounts.
-It is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B.
-Interestingly, nearly all samples with ~38~Sr are located in cluster 2 in model A.
+It is a lithophile element that can substitute for 20Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B.
+Nearly all samples with ~38~Sr are located in cluster 2 in model A.
 
@@
 #| cache: true
-#| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster somewhat based on Fe and Zr content.
+#| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content.
 This results in most of the darker materials clustering towards the bottom of the plot.
 b) Score plot with > 90% Si samples excluded.
-Similar to plot A there is a separation based Fe and Zr values, with darker material towards the bottom of the plot.
-Light-to-colourless samples also remain largely together in cluster 3.
-Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg."
+Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot.
+Light-to-colourless samples remain largely together in cluster 3.
+Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg"
 
@@
 #| out-width: "100%"
-#| fig-cap: "Boxplot of XRF values (%) for model B, grouped by K-means cluster.
-Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with the same (e.g.
+#| fig-cap: "Boxplot of XRF values (%) in PCA model B, grouped by K-means cluster.
+Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with Si (e.g.
 Al, Ca).
-Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content."
+Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content"
 
@@
 
-The K-means cluster analysis of model B is visualised in the NIRS model, which further emphasizes the variation within the general color groups (@fig-nir-xrf-kmeans).
+The K-means cluster analysis of model B is visualised in the NIRS PCA score plot, which further emphasizes the variation within the general colour groups (@fig-nir-xrf-kmeans).
 Cluster 2, with higher ~26~Fe and ~40~Zr values, mainly coincides with the darker material to the right side of the plot.
@@
 They all share a similar hue with D/G peaks detected in the Raman analysis (@fig-nir-sample-comparison).
-All three samples have similar absorbance values, but sample 391 exhibits some differences in the spectra compared to the other two samples.
-Out of the three objects sample 391 seems to stand out more clearly from the others, with the XRF analysis being what mainly seems to separate the other two.
+All three samples have similar absorbance values, but sample 391 exhibits differences in the spectra when compared to the other two samples.
+Samples 386 and 404, however, are separated primarily in the XRF analysis.
 
@@
 The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot.
-This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD."
+This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD"
 
@@
 #| cache: true
-
+#Quarto has issues sizing the figure, perhaps related to the fact that it is made by turning multiple plots into rasters.
+Therefore the base figure import is used here, and not in any other part of the document.
 quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
@@
 
-The exploratory evaluation of spectroscopic data generated on quartz/quartzite points and preforms demonstrates that valuable information on the composition and structural properties of the material can be gained via non-destructive analysis.
-As the identified features are related to the diagenetic/paragenetic environment they should be useful in future survey efforts aimed at locating possible geological sources.
-This does not, however, address the issues related to the resource procurement strategies employed by people within the region during the 4 000 - 2 000 BP period, as the material may have been sourced from other glacial and postglacial landforms, into which the material may have been incorporated.
-One way of addressing this issue would be to construct a predictive model based solely on the spectroscopic fingerprints of the large unknown dataset, which could form the basis of spatial analysis and hypothesis testing.
+The high representation of quartz and quartzite in the Mesolithic - Bronze Age lithic assemblages of Northern Sweden necessitates the development of methods for the characterization of these materials.
+They represent a primary lithic resource, and thus hold a critical position for understanding resource procurement, trade and exchange, and human mobility within the region (as well as other regions [e.g.
+@prieto_etal19]).
+Previous archaeological studies in Northern Sweden have primarily relied on visual characterization, following a system that was devised for managing the large quantities of data generated following the large-scale development of hydroelectric power in the 1940s - 1980s [@biornstad06; @kaeck09: 53-55].
+The division of a heterogeneous material into larger simplified colour groups might, however, give a mistaken impression of homogeneity.
+For example, light grey points with streaks of black mineral running through them might be classified as either light or dark quartzite on different occasions and by different archaeologists.
+The necessity for adopting such classifications in the past is understandable; a certain level of abstraction is necessary in higher level modelling (Sjölander 2022).
+Although this work provided plenty of new insights [e.g.
+@forsberg85; @holm91; @lundberg97], spectroscopic screening provides a more quantitative and empirical approach [@ramacciotti_etal22; @sciuto18].
 
-One benefit of non-destructive screening is the relative ease with which large datasets can be generated.
-Although the resolution may be affected to some degree, it lends itself quite well to the exploratory approach and hypothesis generation.
-A non-destructive approach is quite effective in regions such as Northern Sweden, where little development has been made in regards to the chemical characterization of quartzite and there are many unknowns.
-There is now a better understanding of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as *quartzite*, at least within the Västerbotten area.
-Visually the material varies greatly, both in terms of texture as well as color.
-The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, but as was highlighted in the results, some may have been wrongly classified due to this.
-It is an important reminder to be mindful when aggregating data where the analyst is not the primary producer.
+A benefit of non-destructive screening is the relative ease with which large datasets can be generated.
+Although there are destructive alternatives that provide a higher resolution view of the material’s chemical characteristic,  non-destructive screening lends itself well to exploratory data analysis and hypothesis generation.
+The current study demonstrates the potential of new scientific analysis methods for re-assessing and complementing previous research, and thus increasing our understanding of human resource management of locally available lithic material.
 
-As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former).
-The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus delimiting possible geological source areas.
-Usage of the term "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden.
-Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g.
-@gaweda_cebulak99; @ukar_cloos16].
-Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
+A non-destructive approach is particularly useful in regions with complex geology, such as Northern Sweden, where little development has been made in the chemical characterization of quartzite.
+The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area, with applications in neighbouring regions and beyond.
+The material exhibits considerable variation in terms of texture and colour, and the results show that the initial classification, based on the archaeological documentation, may have included mistakes.
+These mistakes, whilst of little consequence for site interpretation individually, may introduce considerable noise, or “messiness”, when aggregating data for large scale spatio-temporal analyses [@gattiglia15: 2].
 
-Questions of metamorphic grade can also be raised based on the metal-OH features in the NIRS data.
-Should these be related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite).
-Not only does such information provide grounds for discussions on provenance, but it also aids in the construction of a predictive model.
-The main issue, however, seems to be the struggle with sampling dark material using NIRS (and to an extent Raman), which also feature many of the indicators for biotite in the material.
-Sampling in general is a core issue, as the question becomes how representative the results really are of the material/objects in question?
-This does not simply extend to the measurements, however, but also to the visual characterization.
-The division of a heterogeneous material into larger simplified color groups might give a mistaken impression of homogeneity.
-Should a light grey point with streaks of black mineral running through it be classified as a light or dark quartzite?
-The necessity for adopting such classifications in the past is understandable, a certain level of abstraction is necessary in higher level modelling after all [@sjolander22].
-Spectroscopic screening, however, does provide a more quantitative and empirical approach to such tasks [@ramacciotti_etal22].
+The metamorphic nature of quartzite presents a particular challenge for spectroscopic sampling.
+The material frequently exhibits mineral inclusions that produce a varied surface, and the spot size of the probe will affect the user’s ability to capture a representative area of the artefact.
+Likewise, the artefact’s shape, type, and preservation significantly influence the availability of surfaces suitable for sampling.
+Furthermore, dark materials, where a high level of absorbance may saturate measurements or obscure spectral features, are potentially problematic for both NIRS and Raman.
+Although some of these issues may be overcome with other non-destructive methods, such as Hyperspectral Imaging [@sciuto_etal19], the arguably more affordable techniques presented here produce valuable data.
 
-Due to the high representation of quartz and quartzite in the lithic assemblages dating to the Mesolithic - Bronze Age in Northern Sweden [@baudou78], it is vital that methods for characterization of these materials are developed.
-As they represent a primary lithic resource they hold a critical position in the study of resource procurement, trade and exchange and human mobility within the region (as well as many other regions [e.g.
-@prieto_etal19]).
-Previous studies have primarily relied on visual characterization following a system that was setup in order to manage the large quantities of data generated following the large scale development for hydroelectric power during the 1940s - 1980s [@biornstad06; @kaeck09: 53-55].
-This was necessary in order to manage the overabundance of data and make sense of it at the time, and provided a lot of new insights [e.g.
-@forsberg85; @holm91; @lundberg97], but there has been little development in regards to questions of provenance of these materials since.
-This study addresses those needs and demonstrates the potential which exists for increasing the understanding of human resource management of locally available lithic material.
+The spectroscopic features identified in this study relate to the diagenetic/paragenetic environment, and are thus useful for locating potential geological sources in future field surveys.
+As quartzite is a metamorphic rock, there are two factors affecting its chemical and structural properties; the composition of the materials in the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former).
+The D/G bands indicating graphite in the Raman results imply that the material formed in carbon-rich formations.
+Alum shales and similar carbon-rich tectonic strata are present within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], and it would be reasonable to expect carbon structures in quartzite or quartzite-like materials in the region.
+However, the term “graphitic quartzite” is seldom encountered in contemporary Swedish geological publications.
+Usage of the term can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25].
+The occurrence of metamorphic rocks containing graphite, including schist and quartzite, in other regions [e.g.
+@gaweda_cebulak99; @ukar_cloos16] suggests that further studies of this material would be useful for provenance studies in Sweden.
+
+The metal-OH features in the NIRS data can likewise be considered in terms of metamorphism.
+If these are related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite).
+However, this is not possible to confirm without further statistical analyses, such as multi-blocking [@mishra_etal21], integrating the NIRS and XRF data.
+Pursuing this angle further could potentially provide useful information for investigating the material’s provenance.
+
+Whilst spectroscopic methods can provide new data for understanding patterns in resource procurement in geologically complex areas, procurement is only part of the picture for resource management strategies during the Mesolithic – Bronze Age.
+These data can, however, be used to test the different scenarios which existing models represent [e.g.
+@forsberg85; @holm91; @lundberg97].
+For example, a transhumant settlement system in which inland groups, organised along the river valleys, utilize raw materials collected in the mountains, should leave evidence of material flow from the mountains to the forest inland.
+With enough spectroscopic data from known quartzite artefacts, a predictive model could be built to forecast the type of material expected at other sites.
+Multiple hypotheses and models could then be tested, and revised, on the basis of further data from new archaeological excavations and geological exploration.
+The uncertainty of the detailed geology of the region, and the potential noise that glacial disturbance will have introduced, would require that multiple scenarios are tested.
+A more robust geological resource model could be developed by fusing data from multiple types of spectroscopic instrumentation (e.g.
+multiblock PCA; Mishra et al., [-@mishra_etal21]).
 
@@
 
-The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification.
-Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials.
-The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
-Not only is there a need for methodological development in the study of quartzite in an archaeological context, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
-Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.
+This study provides a case study of a non-destructive and exploratory approach to the characterization of petrographic material where reference material is unavailable.
+It provides a step forward in the development of methods for the study of quartzite in an archaeological context, and a template for methodologies that can be applied in regions with complex geology and ambiguous distribution of potential geological sources.
+The results demonstrate clearly that the non-destructive application of all three instrumentations (NIRS, Raman, XRF) is capable of petrographically characterizing some aspect of the quartz and quartzite used in Mesolithic – Bronze Age artefacts in Northern Sweden.
+Used on their own, however, the individual methods lack the resolution needed to provide a fingerprint detailed enough for classification into distinctive groups.
+Nevertheless, the methods do individually provide a more in-depth understanding of the composition and structure of the material in comparison to earlier visual studies, as well as showcase the complexity in the petrographic characters of quartz and quartzite materials.
+The results also demonstrate how the short integration time of these instruments makes it possible to build large datasets that can be statistically explored and mined for information, and ultimately used as the foundation for predictive models for investigating the spatiotemporal aspects of raw material management.
+This would require the integration of archaeological and geological modelling, including theories of resource procurement and economic strategies, with ice sheet and glacial flow dynamics and detailed Quaternary geological mapping.
+
+\linebreak
+
+# Acknowledgements {-}
+
+We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts.
+We would also like to acknowledge cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.
 
@@
 # Author Contributions {-}
-Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft.
+Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft, Writing - Review & Editing.
 
@@
 Paul Geladi: Formal analysis, Methodology
+
+Philip I. Buckland: Writing – original draft, Writing - Review & Editing.
 
@@
 
-# Acknowledgements {-}
-
-We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts.
-We would also like to acknowledge good cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.
+# Data availability
+The code and data used to perform the analysis is openly accessible in an online repository at https://doi.org/10.17605/OSF.IO/F6H98.
 
@@
 \newpage
-
-# Appendix A {#appendix-a}
-
-Collection of loadings for each single instrumentation PCA.
-
-\beginsupplement
-
-\linebreak
-
-```{r}
-#| label: fig-nir-pca-loadings
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.8
-#| out-width: "70%"
-#| fig-cap: "PCA loadings for the first 3 principal components for the NIR model."
-
-quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.jpeg"), error = FALSE)
-
-fig
-```
-
-\linebreak
-
-```{r}
-#| label: fig-raman-pca-loadings
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.8
-#| out-width: "70%"
-#| fig-cap: "PCA loadings for the first 3 principal components for the Raman model."
-
-quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.jpeg"), error = FALSE)
-
-fig
-```
-
-\linebreak
-
-```{r}
-#| label: fig-xrf-pca-loadings-scree
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.8
-#| out-width: "70%"
-#| fig-cap: "PCA Loadings and eigenvalues for the XRF data.
-a) eigenvalues and loadings for the full dataset (model A), b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded (model B)."
-
-fig <- 
-  readRDS(file = here::here("analysis", "data", "derived_data", "009-xrf-loadings.rds"))
-
-fig
-```
-
-\endsupplement
-
-\newpage
-
-#Data availability
-
-The data and code used in the analysis is available at DOI XXX.
 

### 2023-05-16 analysis/paper/paper.qmd
Small corrections to text

@@
 In studies of olivine the absorption region has been determined to be broad and centred around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13].
-Alternatively, weak water features have also been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water.
+Alternatively, weak absorption features (OH) have also been noted in the 950 nm and 1 150 nm bands, particularly among clays.
 However, these overlap with the Fe^2+^ bands in, for example, biotite [@adams75: 104-106].
@@
 \linebreak
-<!--
-```{r fig.width=12, fig.height = 5}
-#| label: fig-nir-spectra-summary
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.8
-#| out-width: "70%"
-#| fig-align: center
-#| fig-cap: "Summary of the raw spectra for each colour group.
-Note the differences in absorbance between the groups, with the dark group featuring a larger range of absorbance values as well as less distinct peaks."
-
-quietly(source(here::here("analysis", "paper", "001-nir-spectra-summary.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.jpeg"), error = FALSE)
-
-fig
-```
-
-\linebreak
-
-```{r fig.width=12, fig.height = 5}
-#| label: fig-nir-savitzky-transform
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.8
-#| out-width: "70%"
-#| fig-align: center
-#| fig-cap: "Comparison of NIR spectra between two samples from the dark group.
-a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering."
-
-quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.jpeg"), error = FALSE)
-
-fig
-```
-
-\linebreak
-
-```{r}
-#| label: fig-nir-savitzky-transform-alt
-#| cache: true
-#| fig-width: 7.87
-#| fig-asp: 0.6
-#| out-width: "70%"
-#| fig-align: center
-#| fig-cap: "Savitzky-Golay transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups."
-
-quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.jpeg"), error = FALSE)
-
-fig
-```
--->
 
@@
 
-The high representation of quartz and quartzite in the Mesolithic - Bronze Age lithic assemblages of Northern Sweden necessitates the development of methods for the characterization of these materials.
+The high representation of quartz and quartzite in the Mesolithic - Bronze Age lithic assemblages of Northern Sweden necessitate the development of methods for the characterization of these materials.
 They represent a primary lithic resource, and thus hold a critical position for understanding resource procurement, trade and exchange, and human mobility within the region (as well as other regions [e.g.

### 2023-05-16 analysis/paper/paper.qmd
Added reference

@@
 ~13~Al and ~19~K is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
-As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism (REF FROM JOHAN), and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
+As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
 Cluster 3 features a more limited range of values between the samples and includes all the brecciated quartz.

### 2023-05-16 analysis/paper/paper.qmd
Updated affiliation

@@
 institute:
-  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
+  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
   - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
@@
 Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights:
-
-* Non-destructive analysis of prehistoric bifacial points made in quartzite using near-infrared spectroscopy, Raman scattering and X-Ray Fluourescence. 
-* Each method detects differences in the structural and chemical characteristics in the material, but on their own the resolution is too low for predictive modelling. 
-* Data fusion techniques are recommended for predictive modelling using non-destructive spectroscopy on quartzite material.
 

### 2023-06-22 analysis/paper/paper.qmd
Updated keywords
Minor manuscript correction

@@
 keywords: |
-  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
+  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
 ---
@@
 @reimann_etal12].
-With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect (REF).
+With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect.
 

### 2023-09-10 analysis/paper/paper.qmd
Added table. Updated to pdf output, referencing tables requires this.

@@
 format:
-  docx:
+  pdf:
+    documentclass: scrartcl
+    fig-pos: "H"
+    geometry: margin = 2.54cm 
     reference-doc: "../templates/template.docx" # Insert path for the DOCX file
@@
 ```{r setup, echo = FALSE}
+library(officer)
+library(officedown)
 knitr::opts_chunk$set(
@@
 \linebreak
- \linebreak
 
 This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite.
-The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map).
+The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data).
 The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south.
@@
 \linebreak
-
+<!--
 ```{r}
@@
 ```
+-->
+
+![Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
+The glacial disturbance is visible in the widely distributed moraine.
+Tectonic symbology after Gee and Stephens (2020).
+Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet](../figures/site_distribution_v1.jpg){#fig-distribution-map width=100%, height=100%}
 
@@
 ```
+-->
+
+![Examples of the bifacial points studied.
+The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
+Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
+Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/figure_2.jpg){#fig-artefacts width=100%, height=100%}
 
 \linebreak
+
+```{r}
+#| label: tbl-point-data
+#| cache: false
+#| cache-lazy: false
+#| tbl-cap: "Table of artefact location, material and type"
+
+suppressPackageStartupMessages(library(flextable))
+
+#custom function for fitting table to page width, see Sarah's answer on
+#https://stackoverflow.com/questions/57175351/flextable-autofit-in-a-rmarkdown-to-word-doc-causes-table-to-go-outside-page-mar
+FitFlextableToPage <- function(ft, pgwidth = 6){
+  ft_out <- ft %>% autofit()
+  ft_out <- width(ft_out, width = dim(ft_out)$widths*pgwidth /(flextable_dim(ft_out)$widths))
+  return(ft_out)
+}
+
+metadata.csv <- 
+  metadata.csv %>% 
+  arrange(site_id)
+
+#set table functions
+set_flextable_defaults(
+  theme_fun = theme_vanilla, 
+  font.size = 10, 
+  padding = 6,
+  background.color = "#FFFFFF")
+
+#columns to select
+columns <- 
+  c("sample_id", "site_id", "latitude", "longitude", "parish", "material", "find_context")
+#names for table header
+names <- 
+  c("Sample ID", "Site", "Latitude", "Longitude", "Parish", "Material", "Find context")
+
+#border options
+inner_border = officer::fp_border(color="gray", width = 1)
+header_border = officer::fp_border(color="black", width = 2)
+body_border = officer::fp_border(color="black", width = 1)
+
+ft <- 
+  #select data
+  flextable(metadata.csv, col_keys = columns) %>% 
+  #set header names
+  set_header_df(mapping = data.frame(keys = columns, values = names, stringsAsFactors = FALSE), key = "keys") %>% 
+  #bold header font
+  bold(bold = TRUE, part = "header") %>% 
+  #align header names center
+  align(align = "center", part = "all") %>% 
+  #set borders and colors 
+  border_inner_h(part = "all", border = inner_border ) %>% 
+  border_inner_v(part = "all", border = inner_border ) %>% 
+  border_outer(part = "body", border = body_border) %>% 
+  border_outer(part = "footer", border = body_border) %>% 
+  border_outer(part = "header", border = header_border) %>% 
+  #use function to fit to page
+  FitFlextableToPage
+
+ft
+```
+
+\pagebreak
 
@@
 
-\linebreak
 \linebreak
@@
 #| fig-asp: 1
-#| out-width: "70%"
+#| out-width: "100%"
 #| fig-align: center
@@
 \linebreak
-
+<!--
+#run this code chunk to display PCA plot without photos
 ```{r}
@@
 #| fig-asp: 1
-#| out-width: "70%"
+#| out-width: "100%"
 #| fig-align: center
 #| fig-cap: "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples"
-#| 
+
 quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
-#knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)
 
@@
 ```
+-->
+
+![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos.jpg){#fig-nir-pca width=100%, height=100%}
 
@@
 #| fig-asp: 0.8
-#| out-width: "70%"
+#| out-width: "100%"
 #| fig-align: center
@@
 \linebreak
-
+<!--
+#run this code chunk to display PCA plot without photos
 ```{r}
 #| label: fig-raman-pca
-#| cache: false
-#| cache-lazy: false
+#| cache: true
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "100%"
+#| fig-align: center
 #| fig-cap: "PCA score plot for Raman data.
@@
 
-#quietly(source(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.jpg"), error = FALSE)
+quietly(source(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8"))
+
+fig
 ```
+-->
+
+![PCA score plot for Raman data.
+Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos.jpg){#fig-raman-pca width=100%, height=100%}
 
@@
 
+<!--
+#run this code chunk to display PCA plot without photos
 ```{r}
@@
 #| cache: true
+#| fig-width: 7.87
+#| fig-asp: 1
+#| out-width: "100%"
+#| fig-align: center
 #| fig-cap: "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content.
@@
 
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.jpg"), error = FALSE)
+quietly(source(here::here("analysis", "paper", "009-xrf-pca.R"), encoding = "UTF-8"))
+
+fig2
 ```
+-->
+
+![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content.
+This results in most of the darker materials clustering towards the bottom of the plot.
+b) Score plot with > 90% Si samples excluded.
+Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot.
+Light-to-colourless samples remain largely together in cluster 3.
+Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos.jpg){#fig-xrf-pca width=100%, height=100%}
 
@@
 #| fig-asp: 0.687
-#| out-width: "70%"
+#| out-width: "100%"
 #| fig-cap: "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points.

### 2023-09-10 analysis/paper/004-nir-pca.R
Updated with fewer material classes.

@@ -62,7 +62,7 @@ fig.1 <-
   labs(x = nir.pc1lab,
        y = nir.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21)) +
+                     values=c(25,21)) +
   scale_fill_manual(name = "Hue",
                     values=pca.colors) +
   guides(fill = guide_legend(override.aes = list(shape = 22,
@@ -77,6 +77,12 @@ fig.1 <-
                               title.position="top",
                               title.hjust = 0.5,
                               order = 2)) +
+  stat_ellipse(aes(color = material, group = material),
+               linetype = 2,
+               lwd = 1.2) +
+  #scale_color_manual(name = "Group",
+  #                   values = c("#36454F", "#E69F00"),
+  #                   limits = c("Dark", "Light")) +
   theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = 1, hjust = 0.02),
         axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
         axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
@@ -95,7 +101,7 @@ fig.2 <-
   labs(x = nir.pc2lab,
        y = nir.pc3lab) +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21)) +
+                     values=c(25,21)) +
   scale_fill_manual(name = "Hue",
                     values=pca.colors) +
   guides(fill = guide_legend(override.aes = list(shape = 22,
@@ -110,6 +116,12 @@ fig.2 <-
                               title.position="top",
                               title.hjust = 0.5,
                               order = 2)) +
+  stat_ellipse(aes(color = hue, group = hue),
+               linetype = 2,
+               lwd = 1.2) +
+  scale_color_manual(name = "Group",
+                     values = c("#36454F", "#E69F00"),
+                     limits = c("Dark", "Light")) +
   theme(plot.title = element_blank(),
         axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
         axis.title.y = element_text(size = 12, face = "bold", colour = "black"),

### 2023-09-10 analysis/paper/007-raman-pca.R

@@ -61,7 +61,7 @@ fig <-
   labs(x = raman.pc1lab,
        y = raman.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21)) +
+                     values=c(25,21)) +
   scale_fill_manual(name = "Hue",
                     values=pca.colors) +
   guides(fill = guide_legend(override.aes = list(shape = 22,

### 2023-09-10 analysis/paper/009-xrf-pca.R

@@ -420,7 +420,7 @@ fig2a <-
                  shape = factor(xrf$material)),
              size = 2) +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21),
+                     values=c(25,21),
                      guide = guide_legend(override.aes = list(fill = "black"),
                                           title.position="top",
                                           title.hjust = 0.5,
@@ -474,7 +474,7 @@ fig2b <-
                  shape = factor(xrf.low$material)),
              size = 2) +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21),
+                     values=c(25,21),
                      guide = "none") +
   scale_color_manual(values = pca.colors,
                      guide = "none") +
@@ -626,7 +626,7 @@ fig4 <-
                     values=c("#0072B2", "#CC79A7","#F0E442"),
                     na.value="#D62728FF") +
   scale_shape_manual(name = "Material",
-                     values=c(25,24,21)) +
+                     values=c(25,21)) +
   guides(fill = guide_legend(override.aes = list(shape = 22,
                                                  fill = c("#0072B2", "#CC79A7","#F0E442"), na.value="#D62728FF",
                                                  ncol = 4,

### 2023-09-10 README.Rmd
Updated readme
Added author

@@
 
-> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal name) <https://doi.org/xxx/xxx>
 
@@
 
-> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal name) `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Sjölander M., Linderholm, J., Geladi P (2023).
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P.
+I.
+(2023).
 (`r format(Sys.Date(), "%Y")`).
@@
 [DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.pdf`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
 

### 2023-09-10 analysis/paper/paper.qmd
Updated manuscript

@@
   dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
+         material == "Brecciated quartz" | material == "Quartzite")
 
@@
   metadata.csv %>% 
+  dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
+                 material == "Brecciated quartz" | material == "Quartzite") %>% 
   arrange(site_id)
@@
 Following log-transformation the data was mean-centred and scaled to unit variance before PCA.
-Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].
+Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].
 
@@
   filter( type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+         material == "Brecciated quartz" | material == "Quartzite") %>% 
   filter(hue == "Dark")
@@
 
-![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos.jpg){#fig-nir-pca width=100%, height=100%}
+![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos_v1.jpg){#fig-nir-pca width=100%, height=100%}
 
@@
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos.jpg){#fig-raman-pca width=100%, height=100%}
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos_v1.jpg){#fig-raman-pca width=100%, height=100%}
 
@@
 Light-to-colourless samples remain largely together in cluster 3.
-Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos.jpg){#fig-xrf-pca width=100%, height=100%}
+Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}
 

### 2023-09-13 README.Rmd
Updated readme

@@
 
-> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal name) <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal of Archaeological Science:Reports) <https://doi.org/xxx/xxx>
 
@@
 
-> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal name) `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal of Archaeological Science:Reports) `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
   - [:file\_folder: paper](/analysis/paper): Quarto source document
-    for manuscript. Includes code to reproduce the figures and tables
+    for manuscript. Includes code in script files to reproduce the figures and tables
     generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
+    `paper.pdf`, suitable for reading (the code is replaced by figures
     and tables in this file)
@@
     illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
+  - [:file\_folder: supplementary-materials](/analysis/supplementary-materials):
     Supplementary materials including notes and other documents

### 2023-09-13 analysis/paper/001-artefact-size.R
Updated scripts

@@ -10,7 +10,7 @@ metadata.csv <-
 Points <-
   metadata.csv %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")  %>%
+                material == "Brecciated quartz" | material == "Quartzite")  %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #plot the length and width

### 2023-09-13 analysis/paper/001-nir-spectra-summary.R

@@ -20,7 +20,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+                material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless")) %>%
   group_by(across(sample_id:weight_g)) %>%
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")

### 2023-09-13 analysis/paper/002-nir-dark-filtered-summary.R

@@ -19,7 +19,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+                material == "Brecciated quartz" | material == "Quartzite") %>%
   dplyr::filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                            "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                            "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>%

### 2023-09-13 analysis/paper/002-nir-peak-summary.R

@@ -19,7 +19,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+                material == "Brecciated quartz" | material == "Quartzite") %>%
   dplyr::filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                                   "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                                   "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>%

### 2023-09-13 analysis/paper/003-nir-savitzky-transform-alt.R

@@ -25,7 +25,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+                material == "Brecciated quartz" | material == "Quartzite") %>%
   filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                            "215","216","229","234","235","237","238","251","262","265","268","272","278","281","282","359","377","392","393","397","405",
                            "406","410","411","413","414","415","416","417","424","425","426","428","430","55","56")) %>%

### 2023-09-13 analysis/paper/003-nir-savitzky-transform.R

@@ -21,7 +21,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless")) %>%
   group_by(across(sample_id:weight_g)) %>%
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")

### 2023-09-13 analysis/paper/004-nir-pca-loadings.R

@@ -22,7 +22,7 @@ nir.merged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                            "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                            "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>%

### 2023-09-13 analysis/paper/004-nir-pca.R

@@ -19,7 +19,7 @@ Points.nir <-
   nir.merged %>%
   filter(
          type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                            "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                            "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>%
@@ -52,9 +52,9 @@ basic_plot2 <-
 
 #bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
 fig.1 <-
-  ggplot(cbind(basic_plot1$data, Points.nir[, c(9,12)]),
+  ggplot(cbind(basic_plot1$data, Points.nir[, c(11,14)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
-  #to add sample id as text
+  #remove hastag to add sample id as text
   #geom_text(aes(label=Points.nir$sample_id, hjust=0.5,vjust=-1.0)) +
   geom_point(size=3.5) +
   theme_bw() +
@@ -77,9 +77,9 @@ fig.1 <-
                               title.position="top",
                               title.hjust = 0.5,
                               order = 2)) +
-  stat_ellipse(aes(color = material, group = material),
-               linetype = 2,
-               lwd = 1.2) +
+  #stat_ellipse(aes(color = hue, group = hue),
+  #             linetype = 2,
+  #             lwd = 1.2) +
   #scale_color_manual(name = "Group",
   #                   values = c("#36454F", "#E69F00"),
   #                   limits = c("Dark", "Light")) +
@@ -91,7 +91,7 @@ fig.1 <-
         legend.position = "bottom")
 
 fig.2 <-
-  ggplot(cbind(basic_plot2$data, Points.nir[, c(9,12)]),
+  ggplot(cbind(basic_plot2$data, Points.nir[, c(11,14)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
   #to add sample id as text
   #geom_text(aes(label=Points.nir$sample_id, hjust=0.5,vjust=-1.0)) +
@@ -116,12 +116,13 @@ fig.2 <-
                               title.position="top",
                               title.hjust = 0.5,
[…]

### 2023-09-13 analysis/paper/005-raman-spectra-summary.R

@@ -22,7 +22,7 @@ raman.baseline.merged <-
 Points.raman <-
   raman.baseline.merged %>%
   filter(type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #Filter raman data to focus on dark samples

### 2023-09-13 analysis/paper/006-raman-peak-identification.R

@@ -25,7 +25,7 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99",
          type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #Filter raman data to focus on dark samples

### 2023-09-13 analysis/paper/007-raman-pca-loadings.R

@@ -27,7 +27,7 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99",
          type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #perform PCA with SNV normalization and mean-center

### 2023-09-13 analysis/paper/007-raman-pca.R

@@ -26,12 +26,12 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99",
          type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #perform PCA with SNV normalization and mean-center
 raman.pca <-
-  prcomp(Points.raman[,c(30:322)], center = TRUE, scale = FALSE)
+  prcomp(Points.raman[,c(32:324)], center = TRUE, scale = FALSE)
 
 #prepare labels for PCs, doing it in 2 steps allows variance percentage to be called on in the Rmd file
 raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
@@ -51,9 +51,9 @@ basic_plot1 <-
 
 #bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
 fig <-
-  ggplot(cbind(basic_plot1$data, Points.raman[, c(9,12)]),
+  ggplot(cbind(basic_plot1$data, Points.raman[, c(11,14)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
-  #to add sample id as text
+  #remove hashtag to add sample id as text
   #geom_text(aes(label=Points.raman$sample_id, hjust=0.5,vjust=-1.0)) +
   geom_point(size=3.5) +
   theme_bw() +

### 2023-09-13 analysis/paper/008-xrf-violin-plot.R

@@ -17,7 +17,7 @@ xrf.merged <-
 Points.xrf <-
   xrf.merged %>%
   filter(type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #Prepare XRF data in long format for violin plots, exclude samples with a dimension smaller than 10 mm

### 2023-09-13 analysis/paper/009-xrf-pca.R

@@ -35,7 +35,7 @@ Points.xrf <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99",
          type == "Point" | type == "Point fragment" | type == "Preform",
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+         material == "Brecciated quartz" | material == "Quartzite") %>%
   replace_na(list(munsell_hue = "Colourless"))
 
 #filter XRF data on sample size (only include samples with smallest dimension >= 10mm),
@@ -51,10 +51,11 @@ xrf <-
 
 #### Impute variables ####
 
+#Listed below are the elements and their listed limit of detection for the Niton XL5 pXRF instrument
 #limit of detection elements Mg, Al, Si, P, S, Cl, K, Ca, V, Cr, Mn, Fe, Co, Cu, Zn, As, Rb, Sr, Y, Zr, Nb, Pd, Ag, Cd, Sn, Sb, Ba, Ti, Pb
 #limit of detection values 2500, 487, 0, 47, 54, 38, 28, 15, 4, 4, 17, 14, 9, 5, 4, 2, 1, 2, 0, 2, 1, 2, 2, 2, 3, 4, 29, 9, 1
 
-#Create vector with limit of detection for variables, this will work as a "censor" of highest possible value
+#create vector with limit of detection for variables, this will work as a "censor" of highest possible value
 dl <- c(487, 0, 28, 15, 14, 2, 9, 29, 2500, 2, 47, 54)
 
 #use log-ratio Expectation - Maximisation algorithm to impute left-censored data (i.e values below LOD)
@@ -71,13 +72,13 @@ xrf <-
 #### PCA full ####
 
 #log-transform data in order to address issues with closure
-#clr can also be used, in that case comment this log10 code, but gives similar results in this instance
+#centered log-ratio can also be used, in that case comment this log10 code; this gives similar results with the current data
 xrf.log <-
   log10(xrf[,5:16])
 
 
 #remove comment from code below if you want to transform data using centered log-ratio
[…]

### 2023-09-13 analysis/paper/010-nir-sample-comparison.R

@@ -45,7 +45,7 @@ raman.averaged <-
 Points.nir <-
   nir.merged %>%
   dplyr::filter(type == "Point" | type == "Point fragment" | type == "Preform",
-                material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
+                material == "Brecciated quartz" | material == "Quartzite")
 
 raman.baseline <-
   read.csv2(here::here("analysis", "data", "derived_data", "raman_baseline_corr.csv"), sep = ",", dec = ".", header = TRUE, check.names = FALSE, na = c("", "NA", "NULL"), encoding = "UTF-8")

### 2023-09-13 analysis/paper/paper.qmd
Updated manuscript

@@
   Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes.
-Provenance of locally available geological material thus becomes complex and fraught with uncertainty.
-Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region.
-However, there is a severe lack of methodological development in the characterization of these materials.
-125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden.
+Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the period 4 000 – 2 000 BP, and is thus critical to the understanding of raw material procurement and material flow within the region.
+However, there is a severe lack of methodological development in the characterization of these materials, and provenance of locally available geological material is complex and fraught with uncertainty.
+126 points and preforms made from quartz/quartzite were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden.
 The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
@@
 Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17].
-This focus has led to a bias, however, which has been highlighted by Prieto et al.
+This focus has led to a bias which has been highlighted by Prieto et al.
 [-@prieto_etal19: 15].
@@
 The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
-Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
+Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].
 
-# Aims {-}
+The geological complexity of both the region and the material necessitates new approaches to studies of both raw materials and artefact distributions.
+This paper presents an assessment of the potential for non-destructive spectroscopy to advance the characterization of quartz and quartzite material without the use of references from known geological sources.
+The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden are limited, however, and thus this aims to:
 
-This paper presents an assessment of the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources.
-The geological complexity of both the region and the material necessitates alternative approaches to studies of raw material management.
-Without relying on well-delimited source areas for geological material, a predictive model capable of separating different “classes” of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models.
-The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden between 4 000 and 2 000 BP.
+* Assess the potential of multispectral non-destructive analysis of quartzite using NIRS, Raman and ED-XRF 
+*	Assess the variability, and identify features, in quartzite that may be of use in its characterization
+*	Discuss the implications of these methods for developing future strategies for studying raw material management of quartzite between 4 000 and 2 000 BP in Northern Sweden
 
@@
 
-This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite.
+This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, brecciated quartz and quartzite.
 The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data).
 The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south.
-The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.
+The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and ED-XRF (@fig-artefacts; @fig-artefact-size).
 
-The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89].
+The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made for Northern Sweden [@baudou92: 99; @forsberg85: 5; @forsberg89].
 The latter suggest a period of use between ca.
-3 800 - 2 000 BP [@kaeck09: 60].
+3 800 – 2 000 BP [@kaeck09: 60].
 Studies of raw material use at sites dating to this period seem to indicate an increase in the use of quartzite when compared to earlier periods [@baudou78].
@@
 \linebreak
-<!--
-```{r}
-#| label: fig-distribution-map
-#| cache: false
-#| cache-lazy: false
-#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
-The glacial disturbance is visible in the widely distributed moraine.
-Tectonic symbology after [@gee_stephens20].
-Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"
 
-knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
-```
--->
-
-![Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
-The glacial disturbance is visible in the widely distributed moraine.
-Tectonic symbology after Gee and Stephens (2020).
-Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet](../figures/site_distribution_v1.jpg){#fig-distribution-map width=100%, height=100%}
-
-\linebreak
-
-The dataset includes three main groups of petrographic material, as classified by archaeologists: quartz, quartzite, and brecciated quartz (Swe.
-"*brecciekvarts*").
-Artefacts that had no prior classification were sorted into either the quartz or quartzite category based on visual inspection.
-This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples.
-The inclusion of quartz and brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
-
-In his analysis of the petrographic material in the Early Norrland project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
-Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis.
-*Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
-@holm91: 24].
-As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.
-
-\linebreak
-<!--
-```{r}
-#| label: fig-artefacts
-#| cache: false
-#| cache-lazy: false
-#| fig-cap: "Examples of the bifacial points studied.
-The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
-Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
-Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"
-
-knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
-```
--->
-
-![Examples of the bifacial points studied.
-The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
-Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
-Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/figure_2.jpg){#fig-artefacts width=100%, height=100%}
+![Geographical distribution of the sampled sites, with find context, along the river of Ångermanälven, Västerbotten, Sweden.
+The Late Quaternary reincorporation of geological material from the Scandian mountains into the sediments in the forest inlands adds to the complexity of questions of provenance.
+Tectonic symbology after (Gee and Stephens 2020).
+Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.](../figures/Fig1.jpg){#fig-distribution-map width=100%, height=100%}
 
@@
 #| cache-lazy: false
-#| tbl-cap: "Table of artefact location, material and type"
+#| tbl-cap: "List of artefact location, material and type."
 
@@
 
-\pagebreak
+\linebreak
 
-# Methods {-}
+The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and brecciated quartz (Swe.
+"*brecciekvarts*") (@fig-artefacts; @tbl-point-data).
+Geological definitions of quartzite include characteristics such as the material being exceptionally hard, predominantly consisting of quartz, and with conchoidal fracturing [@skolnick65; @howard05; @prieto_etal19].
+A reliable classification would thus be based on a combination of chemical and physical (macroscopic and microscopic) properties.
+Archaeologists, however, tend to classify the material in the field on the basis of visual characteristics and perceived hardness.
+This can lead to considerable ambiguity, but in general, a material will be archaeologically classified as quartzite in Northern Sweden if it is an exceptionally hard, fine-grained, quartz-rich rock.
 
 \linebreak
+
+![Examples of the bifacial points studied.
+The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
+Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
+Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/Fig2.jpg){#fig-artefacts width=100%, height=100%}
+
 \linebreak
 
-All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/).
-The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens.
-The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data).
-The artefacts range from small and flat point fragments, to large and rough preforms.
-Therefore, the signal integration could at times be difficult, leading to scattering effects and saturation.
-For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult.
-When sampling the probe was in contact with the surface of the artefact at as flat a surface as possible.
-If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g.
-the dark mineral veins in @fig-artefact-size).
-
-For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object.
-The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution.
-The risk of misinterpretation is, however, somewhat compensated by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source.
-The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.
-
-In colour, the artefacts exhibit a mainly black or grey hue, with a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
-While the colour of the material has been classified according to the Munsell system (see supplementary data), for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
-In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.
+This study includes `r length(quartzite)` and `r length(bquartz)` brecciated quartz samples.
+The inclusion of brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
+The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67].
+Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis.
+No detailed information on these microscopic features were mentioned in his original publication.
+Brecciated quartz in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g.
+@holm91: 24] (@fig-artefacts).
+As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.
 
@@
 
+# Methods {-}
+
+\linebreak
+ \linebreak
+
+All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/).
+The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens.
+The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data).
+The artefacts range from small and flat point fragments, to large and rough preforms.
+This variation led to scattering effects and saturation which at times made signal integration difficult.
+
+For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult.
+When sampling, the probe was positioned in contact with as flat a surface as possible of the artefact.
+If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g.
+the dark mineral veins in @fig-artefact-size).
+
+For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object.
+The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution.
+The risk of misinterpretation is, however, somewhat compensated for by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source.
+The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.
+
+In colour, the artefacts exhibit a mainly black or grey hue, with a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
+While the colour of the material has been classified according to the Munsell system (see supplementary data), for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
+In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.
+
+\linebreak
+
 ## Instrumentation and data pre-processing {-}
@@
 
-Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near-infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
-The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify it.
-Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
-The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].
-Overtones occur at frequencies that are a multiple of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].
+On the basis of the theory of molecular vibration and excitation [@hunt77], it is possible to generate overtone vibrations when exposing a geological material to electromagnetic radiation.
+These overtones can be related to certain molecular groups, the hydroxyl (OH) group being the most common one capable of generating them.
+In quartz-based materials this typically relates to molecular water and metal-OH combinations [@hunt77: 508].
+Overtones occur at frequencies that are multiples of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].
 
 NIR analysis was conducted with an Analytical Spectral Device (ASD) LabSpec 4, using a contact probe (spot size: 10 mm).
-The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp.
-spectral resolution) of 1.4 nm (resp.
-3 nm) in the visible and near infrared range and 1.1 nm (resp.
-10 nm) in the short‐wave infrared range.
+The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling of 1.4 nm in the visible and near-infrared range and 1.1 nm in the short‐wave infrared range.
+Respectively, spectral resolution is 3 nm in the visible and near-infrared range, and 10 nm in the short-wave infrared range.
 Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
@@
 ### Raman {-}
-Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared region [@nafie01].
-Raman spectroscopy is based on the *Raman scattering* effect.
+Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structures based on their vibrational properties within the near-infrared region [@nafie01].
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
-The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12].
+The subsequent light that is scattered contains both *Rayleigh* and *Raman scatter*, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12].
 The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency.
-As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].
+As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material[@smith_carabatos-nedelec01].
 
-The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
+The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser (spot size: 85 $\mu$m).
 The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^.
@@
 ### X-Ray Fluorescence (XRF) {-}
-XRF spectroscopy is a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
-Spectral features in XRF spectrometry can be related to certain elements according to Moseley’s law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24].
-Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159].
-In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample.
-One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.
-XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand.
+XRF spectroscopy is a well-established method for petrographic characterization in archaeology [@hall60; @shackley11a].
+In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample [@sitko_zawisa12: 143-159].
+One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated, as was done in this project.
+
+Measurements were collected using the Energy Dispersive (ED) Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand.
 The reference calibration used for element quantification was mining mode [@knight_etal21: 2].
 The integration time was set to 130 seconds.
+Higher spectral resolution can be achieved if used with a Wavelength Dispersive (WD) XRF system [@menne_etal20].
+However, these are limited to lab-based environments and not applicable in the field, which is central to the current study.
 
 As the variables in XRF analysis are given as proportions of a constant sum (e.g.
-percentage, parts per million (ppm)), this creates something referred to as the closed sum effect [@pawlowsky-glahn_egozcue06].
+percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06].
 This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20].
-This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99].
-Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15].
+This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99].Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15].
 In the PCA, the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
@@
 Following log-transformation the data was mean-centred and scaled to unit variance before PCA.
-Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].
+To prevent quantification errors related to samples size artefacts that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].
 
@@
 The data were evaluated using PCA, an unsupervised exploratory technique for reducing the dimensionality of a dataset [@geladi_linderholm20].
-Multivariate approaches, such as PCA, have been proven useful for exploring complex spectroscopic datasets [@sciuto18].
-Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material.
-These groupings can then be followed up with supervised methods, and aid in the construction of predictive models for classifying unknown materials according to previously defined groups.
-
-PCA analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].
+Analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].
 
@@
 Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm.
-These features are all the result of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19].
+These features are all the result of overtones and vibration combinations of OH groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water  (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19].
 OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region.
-While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH [@hunt77: 509; @clark_etal90: 12664].
+While there are exceptions, a more intense 2 200 nm band has typically been related to aluminium hydroxide (AlOH), and a more intense 2 350 nm band to magnesium hydroxide (MgOH) [@hunt77: 509; @clark_etal90: 12664].
 In the current dataset the 2 200 nm band is typically the more intense band.
@@
 This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative.
-A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %).
-Objects with a white or light grey hue gravitate towards the bottom left of the score plot, while material that displays some translucence cluster towards the top left.
+A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %).Objects with a white or light grey hue gravitate towards the bottom left of the score plot, while material that displays some translucence cluster towards the top left.
 There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca :A.
@@
 There are two distinguishing features present in the Raman data (@fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
-The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, , are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
+The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
 
-The second feature shows as two peaks at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominates the spectra in terms of intensity for the darker samples.
+second feature shows as two peaks at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominates the spectra in terms of intensity for the darker samples.
 These positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99].
-The D band occurs as a result of disordered carbon (C) structures, whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
+The D band occurs as a result of disordered C structures, whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
 The G band peak falls within the expected position, whereas the position of the D band is more commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
@@
 that is obscured by the D band.
-Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been identified as one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
+Two peaks can also be found in the 700 - 800 cm^-1^ band range.
+The 700 cm^-1^ band has been identified as one of the aforementioned quartz bands.
+Studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
 
@@
 #| fig-align: center
-#| fig-cap: "TAveraged and baseline corrected Raman spectra for the light and dark groups.
+#| fig-cap: "Averaged and baseline corrected Raman spectra for the light and dark groups.
 The features can be related to quartz and carbon structures in the material.
-The initial three peaks in 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
-The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures"
+The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
+The 1 287 cm^-1^ and 1 598 cm^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures.
+The white and colourless spectra only show significant peaks for quartz (see Supplementary Data)."
 
@@
 It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals.
-It does not fit with the “expected” image of a quartzite and may be a case of erroneous classification.
+It does not fit with the “expected” image of a quartzite and may be a case of erroneous archaeological classification.
 The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.
@@
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster"
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster."
 
@@
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos_v1.jpg){#fig-raman-pca width=100%, height=100%}
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.](../figures/007-raman-pca-with-photos_v1.jpg){#fig-raman-pca width=100%, height=100%}
 
@@
 The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18].
-The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot).
+The analysis shows that ~14~Si makes up the main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot).
 The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
@@
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
-~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
+~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn) recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
 
@@
 Of particular note is the bimodal distribution of Si values.
-Y and Zn were left out as they had less than 30 values above LOD"
+Y and Zn were left out as they had less than 30 values above LOD."
 
@@
 Cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
-There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values.
+There is no real pattern to this cluster aside from higher ~14~Si values with lower iron (~26~Fe) and zirconium (~40~Zr) values.
 Loadings for elements expected to follow changes in 14Si content also end up near cluster 1 (e.g.
-Al, Ca).
-One of the “quartz” labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not display any typical quartz characteristics.
+Al, calcium (Ca)).
 
-Cluster 2 is distinguished by a higher ~22~Ti, ~26~Fe and ~40~Zr content.
+Cluster 2 is distinguished by a higher titanium (~22~Ti), ~26~Fe and ~40~Zr content.
 It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue.
-~13~Al and ~19~K is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
+~13~Al and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
 As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
@@
 The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot.
-This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD"
+This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD."
 
@@
 ![Comparison of three dark samples from separate K-cluster groups (model B).
-Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples](../figures/010-nir-sample-comparison.jpeg){#fig-nir-sample-comparison width=7.87in, height=7.87in}
+Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples.](../figures/010-nir-sample-comparison.jpeg){#fig-nir-sample-comparison width=7.87in, height=7.87in}
 
@@
 
-The high representation of quartz and quartzite in the Mesolithic - Bronze Age lithic assemblages of Northern Sweden necessitate the development of methods for the characterization of these materials.
-They represent a primary lithic resource, and thus hold a critical position for understanding resource procurement, trade and exchange, and human mobility within the region (as well as other regions [e.g.
-@prieto_etal19]).
-Previous archaeological studies in Northern Sweden have primarily relied on visual characterization, following a system that was devised for managing the large quantities of data generated following the large-scale development of hydroelectric power in the 1940s - 1980s [@biornstad06; @kaeck09: 53-55].
-The division of a heterogeneous material into larger simplified colour groups might, however, give a mistaken impression of homogeneity.
-For example, light grey points with streaks of black mineral running through them might be classified as either light or dark quartzite on different occasions and by different archaeologists.
-The necessity for adopting such classifications in the past is understandable; a certain level of abstraction is necessary in higher level modelling (Sjölander 2022).
-Although this work provided plenty of new insights [e.g.
-@forsberg85; @holm91; @lundberg97], spectroscopic screening provides a more quantitative and empirical approach [@ramacciotti_etal22; @sciuto18].
+A screening-based approach to material analysis using non-destructive spectroscopy is particularly useful in regions with complex geology, such as Northern Sweden, where little development has been made in the chemical characterization of quartzite.
+With so many unknowns, it is better to first assess the variability within the material before engaging in more detailed and targeted analysis.
+Non-destructive methods are capable of generating large and expansive datasets with relative ease, and little risk to the material in question.
+When coupled with exploratory data analysis, these techniques are a useful starting point for an empirically grounded description of a material and for identifying any features of note.
+Where feasible, the material can then be further analysed with destructive methods capable of providing a higher resolution view of chemical and physical characteristic, potentially leading to its general characterisation [e.g.
+@blomme_etal12; @prieto_etal19; @prieto_etal20].
 
-A benefit of non-destructive screening is the relative ease with which large datasets can be generated.
-Although there are destructive alternatives that provide a higher resolution view of the material’s chemical characteristic,  non-destructive screening lends itself well to exploratory data analysis and hypothesis generation.
-The current study demonstrates the potential of new scientific analysis methods for re-assessing and complementing previous research, and thus increasing our understanding of human resource management of locally available lithic material.
+The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area.
+These results are applicable to neighbouring regions and beyond, and should help in furthering research into prehistoric tools in Scandinavia and other regions.
+The material exhibits considerable variation in terms of texture and colour, and the results suggest that the initial classification, based on the archaeological documentation, may have included mistakes.
+These mistakes, whilst of little consequence for site interpretation individually, may introduce considerable noise, or “messiness”, when aggregating data for large scale spatio-temporal analyses [@gattiglia15: 2].
+The results also highlight some of the implications that the choice of spectroscopic analysis method may have for the identification of features in a material.
+In particular, they demonstrate that reliance on any individual method may introduce limitations which could make material classification difficult or misleading.
 
-A non-destructive approach is particularly useful in regions with complex geology, such as Northern Sweden, where little development has been made in the chemical characterization of quartzite.
-The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area, with applications in neighbouring regions and beyond.
-The material exhibits considerable variation in terms of texture and colour, and the results show that the initial classification, based on the archaeological documentation, may have included mistakes.
-These mistakes, whilst of little consequence for site interpretation individually, may introduce considerable noise, or “messiness”, when aggregating data for large scale spatio-temporal analyses [@gattiglia15: 2].
+The strength of this study is in the multispectral approach; the data generated from each method providing a complementary image of the structural and compositional nature of the material.
+Although Raman spectroscopy generated a limited amount of information in comparison with the other methods, the graphite structures identified in the material will likely be significant in future provenance studies.
+The utility of Raman spectroscopy as a screening tool, and in comparison to NIRS and XRF, seems to be somewhat limited when applied to quartz-based materials.
+Similarly, the broad bands in NIRS spectra are complex and can be difficult to interpret.
+Nevertheless, insights from the metal-OH bands suggest that more information on metamorphic grades may be obtainable.
+If the bands are related to the Al content of the material and the presence of mica, then the band positions may indicate differing metamorphic grades in the material (muscovite - biotite).
+This will need to be evaluated in more detail in future studies.
+
+Whilst NIRS and Raman may be somewhat limited in the characterization of quartz and quartzite they have considerable potential when combined with the XRF.
+The complementary information about the structural elements in the material can help guide interpretation of the compositional data.
+Ideally these different sources would be fused using methods such as multi-block PCA to better asses key variables contributing to the model [@mishra_etal21].
 
@@
 Likewise, the artefact’s shape, type, and preservation significantly influence the availability of surfaces suitable for sampling.
-Furthermore, dark materials, where a high level of absorbance may saturate measurements or obscure spectral features, are potentially problematic for both NIRS and Raman.
+This can be further compounded by certain limitations with the technique itself.
+Both NIRS and Raman can be affected by saturation on darker materials, due in part to the absorbance, obscuring spectral features.
 Although some of these issues may be overcome with other non-destructive methods, such as Hyperspectral Imaging [@sciuto_etal19], the arguably more affordable techniques presented here produce valuable data.
+Not everyone will have the benefit of working with all three instrumentations however, and in such situations XRF seems to provide more utility.
+Although WD-XRF systems will provide higher spectral resolution and element detection [@menne_etal20], portable ED-XRF has come a long way since its introduction, and instruments becoming commonly available in archaeological laboratories.
 
-The spectroscopic features identified in this study relate to the diagenetic/paragenetic environment, and are thus useful for locating potential geological sources in future field surveys.
+The spectroscopic features identified in this study relate to the diagenetic/paragenetic environment and are thus useful for locating potential geological sources in future field surveys.
 As quartzite is a metamorphic rock, there are two factors affecting its chemical and structural properties; the composition of the materials in the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former).
 The D/G bands indicating graphite in the Raman results imply that the material formed in carbon-rich formations.
-Alum shales and similar carbon-rich tectonic strata are present within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], and it would be reasonable to expect carbon structures in quartzite or quartzite-like materials in the region.
+Alum shales and similar carbon-rich tectonic strata are present within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], and it would be reasonable to expect C structures in quartzite or quartzite-like materials in the region.
 However, the term “graphitic quartzite” is seldom encountered in contemporary Swedish geological publications.
@@
 
-The metal-OH features in the NIRS data can likewise be considered in terms of metamorphism.
-If these are related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite).
-However, this is not possible to confirm without further statistical analyses, such as multi-blocking [@mishra_etal21], integrating the NIRS and XRF data.
-Pursuing this angle further could potentially provide useful information for investigating the material’s provenance.
-
-Whilst spectroscopic methods can provide new data for understanding patterns in resource procurement in geologically complex areas, procurement is only part of the picture for resource management strategies during the Mesolithic – Bronze Age.
+The high representation of quartz and quartzite in the Mesolithic - Bronze Age lithic assemblages of Northern Sweden necessitates the development of methods for the characterization of these materials.
+They represent a primary lithic resource, and thus hold a critical position for understanding resource procurement, trade and exchange, and human mobility within the region (as well as other regions [e.g.
+@prieto_etal19].
+Whilst spectroscopic methods can provide new data for advancing the understanding of patterns in resource procurement in geologically complex areas, procurement is only part of the picture for resource management strategies during the Mesolithic – Bronze Age.
 These data can, however, be used to test the different scenarios which existing models represent [e.g.
-@forsberg85; @holm91; @lundberg97].
+@forsberg85; @bergman95; @lundberg97].
 For example, a transhumant settlement system in which inland groups, organised along the river valleys, utilize raw materials collected in the mountains, should leave evidence of material flow from the mountains to the forest inland.
-With enough spectroscopic data from known quartzite artefacts, a predictive model could be built to forecast the type of material expected at other sites.
+With high quality spectroscopic data from existing quartzite artefacts, a predictive model could be built to forecast the type of material expected at other sites.
 Multiple hypotheses and models could then be tested, and revised, on the basis of further data from new archaeological excavations and geological exploration.
 The uncertainty of the detailed geology of the region, and the potential noise that glacial disturbance will have introduced, would require that multiple scenarios are tested.
-A more robust geological resource model could be developed by fusing data from multiple types of spectroscopic instrumentation (e.g.
-multiblock PCA; Mishra et al., [-@mishra_etal21]).
+A more robust geological resource model could be developed by fusing data from multiple types of spectroscopic instrumentation [e.g.
+multiblock PCA; @mishra_etal21].
 
@@
 
-This study provides a case study of a non-destructive and exploratory approach to the characterization of petrographic material where reference material is unavailable.
-It provides a step forward in the development of methods for the study of quartzite in an archaeological context, and a template for methodologies that can be applied in regions with complex geology and ambiguous distribution of potential geological sources.
-The results demonstrate clearly that the non-destructive application of all three instrumentations (NIRS, Raman, XRF) is capable of petrographically characterizing some aspect of the quartz and quartzite used in Mesolithic – Bronze Age artefacts in Northern Sweden.
-Used on their own, however, the individual methods lack the resolution needed to provide a fingerprint detailed enough for classification into distinctive groups.
-Nevertheless, the methods do individually provide a more in-depth understanding of the composition and structure of the material in comparison to earlier visual studies, as well as showcase the complexity in the petrographic characters of quartz and quartzite materials.
-The results also demonstrate how the short integration time of these instruments makes it possible to build large datasets that can be statistically explored and mined for information, and ultimately used as the foundation for predictive models for investigating the spatiotemporal aspects of raw material management.
-This would require the integration of archaeological and geological modelling, including theories of resource procurement and economic strategies, with ice sheet and glacial flow dynamics and detailed Quaternary geological mapping.
+This study provides, through a non-destructive multispectral approach, empirical insights into the variability of quartzite material in Northern Sweden.
+The development of non-destructive screening methods provides a good initial basis for future detailed studies of quartz and quartzite material.
+This is especially valuable in geologically complex regions where there is an absence of reference materials and ambiguous distribution of potential geological sources.
+Through the use of multiple instrumentations, the study has identified a number of features in quartzite that may be relevant to its future characterization.
+These include different forms of mica (NIRS), graphite (Raman), and iron content (ED-XRF).
+When coupled with exploratory data analysis, and with the addition of further geological and archaeological data, it should be possible to generate hypotheses on the flow of these materials in the Mesolithic – Bronze Age of the region.
+These hypotheses could then be tested through the analysis of material from new excavations and surveys.
+Targeted studies using destructive methods (e.g.
+thin sections analysis, laser ablation), capable of providing more detailed analysis and potentially a classification of the material, could be strategically employed to increase the robustness of the study.
+In order to truly advance the topic of prehistoric raw material management in Northern Sweden, it will be necessary to combine such studies with both archaeological and geological models.
+These will need to integrate archaeological and geological survey data with theories of resource procurement and economic strategies, ice sheet and glacial flow dynamics and detailed Quaternary geological mapping.
 

### 2023-09-13 analysis/paper/paper.qmd
Updated manuscript

@@
 
-In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden.
+In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden (https://samlingar.vbm.se/).
 X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data.
@@
 
-This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, brecciated quartz and quartzite.
+This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, *brecciated quartz* and quartzite.
 The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data).
@@
 
-The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and brecciated quartz (Swe.
+The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and *brecciated quartz* (Swe.
 "*brecciekvarts*") (@fig-artefacts; @tbl-point-data).
@@
 
-This study includes `r length(quartzite)` and `r length(bquartz)` brecciated quartz samples.
-The inclusion of brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
+This study includes `r length(quartzite)` and `r length(bquartz)` *brecciated quartz* samples.
+The inclusion of *brecciated quartz* in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
 The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67].
@@
 No detailed information on these microscopic features were mentioned in his original publication.
-Brecciated quartz in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g.
+*Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g.
 @holm91: 24] (@fig-artefacts).
@@
 In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample [@sitko_zawisa12: 143-159].
-One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated, as was done in this project.
+Portable XRF instruments typically feature a set of filters to achieve this, calibrated for different mediums (e.g.
+soil, metals, minerals) [@knight_etal21: 2].
 
 Measurements were collected using the Energy Dispersive (ED) Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand.
-The reference calibration used for element quantification was mining mode [@knight_etal21: 2].
+The reference calibration used for element quantification was mining mode [@knight_etal21: 2; @thermo18].
 The integration time was set to 130 seconds.
@@
 Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot.
-One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material.
+One of these is a *brecciated quartz* sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material.
 The second sample, however, stands out from the rest of the material.
@@
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
-~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn) recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
+Certain elements recorded a high number of values below the LOD (“missing values”), including ~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn).
+Although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
 
@@
 As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
-Cluster 3 features a more limited range of values between the samples and includes all the brecciated quartz.
+Cluster 3 features a more limited range of values between the samples and includes all the *brecciated quartz*.
 

### 2024-02-07 analysis/paper/paper.qmd
Small edit
Review corrections

@@
 abstract: |
-  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes.
+  Northern Fennoscandia is a geologically complex region affected by both glacial and postglacial processes.
 Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the period 4 000 – 2 000 BP, and is thus critical to the understanding of raw material procurement and material flow within the region.
 However, there is a severe lack of methodological development in the characterization of these materials, and provenance of locally available geological material is complex and fraught with uncertainty.
-126 points and preforms made from quartz/quartzite were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden.
+126 quartz/quartzite points and preforms were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden.
 The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
@@
 Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17].
-This focus has led to a bias which has been highlighted by Prieto et al.
+This focus has led to a material bias which has been highlighted by Prieto et al.
 [-@prieto_etal19: 15].
-Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @columbu_etal23; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19; @ramacciotti_etal22; @ramacciotti_etal23; @sherman_etal23].
 
@@
 There has, however, been little development in the petrographic characterization and provenance study of quartz and quartzite [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-).
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
@@
 The Late Quaternary reincorporation of geological material from the Scandian mountains into the sediments in the forest inlands adds to the complexity of questions of provenance.
-Tectonic symbology after (Gee and Stephens 2020).
+Geological symbology after (Gee and Stephens 2020).
 Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.](../figures/Fig1.jpg){#fig-distribution-map width=100%, height=100%}
@@
 
-This study includes `r length(quartzite)` and `r length(bquartz)` *brecciated quartz* samples.
+This study includes `r length(quartzite)` quartzite and `r length(bquartz)` *brecciated quartz* samples.
 The inclusion of *brecciated quartz* in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.
@@
 These features are all the result of overtones and vibration combinations of OH groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water  (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19].
-OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region.
+Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region.
 While there are exceptions, a more intense 2 200 nm band has typically been related to aluminium hydroxide (AlOH), and a more intense 2 350 nm band to magnesium hydroxide (MgOH) [@hunt77: 509; @clark_etal90: 12664].
@@
 The 2 200 nm and 2 350 nm absorption bands can be found in mica, such as muscovite and biotite, which can be found in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207].
-Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2 250 nm band for biotite [@duke94].
+Although muscovite and biotite share some band positions, the altered aluminium (Al) content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2 250 nm band for biotite [@duke94].
 As can be seen in the average spectra in @fig-nir-spectra-summary, the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.
 
-Two smaller features were recorded, albeit vague and difficult to make out in the non-processed spectra, at 1 065 nm and 1 300 nm (@fig-nir-spectra-summary).
+Two smaller features were recorded, albeit vague and difficult to make out in the non-processed spectra, at 1 065 nm and 1 300 nm (@fig-nir-peak-summary).
 These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
@@
 The 700 cm^-1^ band has been identified as one of the aforementioned quartz bands.
-Studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
+Studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to silicon (Si), aluminium (Al) and hydroxyl (OH) vibrations within the 700 - 800 cm^-1^ region.
 
@@
 The XRF spectrometer is not capable of detecting all the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
-The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18].
-The analysis shows that ~14~Si makes up the main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (@fig-xrf-violin-plot).
-The majority of the readings resulted in ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
+The elements represented in this group are all those with an atomic number lower than magnesium (~12~Mg) [@thermo18].
+The analysis shows that silicon (~14~Si) makes up the main bulk of the material, with aluminium (~13~Al) recorded up to ~ 5% for some samples (@fig-xrf-violin-plot).
+The majority of the readings resulted in silicon (~14~Si) content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
 Whilst this could be due to the limitations of the instrumentation, it could also be a result of the sampling.
 As quartzite is a metamorphic rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement.
-Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
-Certain elements recorded a high number of values below the LOD (“missing values”), including ~12~Mg, vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn).
-Although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
+Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a silicon (~14~Si) content higher than 90%.
+Certain elements recorded a high number of values below the LOD (“missing values”), including magnesium (~12~Mg), vanadium (~23~V), manganese (~25~Mn), yttrium (~39~Y) and zinc (~30~Zn).
+Although in the case of magnesium (~12~Mg) and manganese (~25~Mn) the recorded values reached up to 0.3% and 3% respectively.
 
-Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values.
-Instead log10-transformation was used, producing a similar model, but one that managed the loading for ~14~Si better.
-The PCA was run twice, firstly with the > 90% ~14~Si samples included and secondly excluded.
+Initial analysis used clr transformed data, but encountered issues related to the structure of the silicon (~14~Si) values.
+Instead log10-transformation was used, producing a similar model, but one that managed the loading for silicon (~14~Si) better.
+The PCA was run twice, firstly with the > 90% silicon (~14~Si) samples included and secondly excluded.
 The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
@@
 As can be seen in the score plot (@fig-xrf-pca :A), the three K-means clusters remain separated.
-Cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
-There is no real pattern to this cluster aside from higher ~14~Si values with lower iron (~26~Fe) and zirconium (~40~Zr) values.
-Loadings for elements expected to follow changes in 14Si content also end up near cluster 1 (e.g.
-Al, calcium (Ca)).
+Cluster 1 is made up of those samples that feature a high silicon (~14~Si) content (> 90%).
+There is no real pattern to this cluster aside from higher silicon (~14~Si) values with lower iron (~26~Fe) and zirconium (~40~Zr) values.
+Loadings for elements expected to follow changes in silicon( ~14~Si) content also end up near cluster 1 (e.g.
+aluminium (~13~Al), calcium (~20~Ca)).
 
-Cluster 2 is distinguished by a higher titanium (~22~Ti), ~26~Fe and ~40~Zr content.
+Cluster 2 is distinguished by a higher titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) content.
 It mainly features samples with a dark hue, with the few samples from the light group being grey in colour and arguably somewhere in between a “dark” or “light” hue.
-~13~Al and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
-As ~13~Al and ~19~K are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
+Aluminium (~13~Al) and potassium (~19~K) is similarly higher in cluster 2 than cluster 3, which is in turn marked by its, on average, lower elemental content than the other two clusters.
+As aluminium (~13~Al) and potassium (~19~K) are indicative of a sedimentary depositional context that has been subject to metamorphism [@hazen_morrison2022], and the cluster with higher values may thus indicate samples with a different grade of metamorphism.
 Cluster 3 features a more limited range of values between the samples and includes all the *brecciated quartz*.
@@
 The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples.
-Cluster 2 features the main bulk of darker samples with higher  ~22~Ti, ~26~Fe and ~40~Zr values.
-Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values.
-It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.
+Cluster 2 features the main bulk of darker samples with higher titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) values.
+Cluster 1 likewise feature a large portion of the darker materials, with comparable titanium (~22~Ti), iron (~26~Fe) and zirconium (~40~Zr) values.
+It also has higher values of magnesium (~12~Mg), aluminium (~13~Al), sulphur (~16~S), potassium (~19~K), and calcium (~20~Ca).
 
-~38~Sr was detected at trace amounts.
-It is a lithophile element that can substitute for 20Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B.
-Nearly all samples with ~38~Sr are located in cluster 2 in model A.
+Strontium (~38~Sr) was detected at trace amounts.
+It is a lithophile element that can substitute for calcium (~20~Ca) in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B.
+Nearly all samples with strontium (~38~Sr) are located in cluster 2 in model A.
 
@@
 b) Score plot with > 90% Si samples excluded.
-Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot.
+Similar to plot A there is a separation based on ~26~Fe and ~40~Zr values, with darker material towards the bottom of the plot.
 Light-to-colourless samples remain largely together in cluster 3.
-Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}
+Cluster 1 stands out by featuring slightly lower ~14~Si, ~26~Fe, ~22~Ti, and ~40~Zr values, but higher values in ~13~Al, ~20~Ca, ~19~K, and ~12~Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}
 
@@
 #| fig-cap: "Boxplot of XRF values (%) in PCA model B, grouped by K-means cluster.
-Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with Si (e.g.
-Al, Ca).
-Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content"
+Cluster 1 is made up of samples with lower values of ~14~Si, but higher values for those elements that are typically associated with Si (e.g.
+~13~Al, ~20~Ca).
+Cluster 2 includes the samples with higher ~26~Fe/~22~Ti/~40~Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low ~26~Fe/~22~Ti/~40~Zr content"
 
@@
 The K-means cluster analysis of model B is visualised in the NIRS PCA score plot, which further emphasizes the variation within the general colour groups (@fig-nir-xrf-kmeans).
-Cluster 2, with higher ~26~Fe and ~40~Zr values, mainly coincides with the darker material to the right side of the plot.
+Cluster 2, with higher iron (~26~Fe) and zirconium (~40~Zr) values, mainly coincides with the darker material to the right side of the plot.
 This was expected as the trend along PC 1 in the NIR PCA mainly separates the material according to absorbance.
@@
 The results presented here have increased the state of knowledge of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as quartzite within the Västerbotten area.
-These results are applicable to neighbouring regions and beyond, and should help in furthering research into prehistoric tools in Scandinavia and other regions.
+This non-destructive exploratory approach is particularly useful in regions where there is a lack of data and an element of uncertainty (see Jebb et al., [-@jebb_etal17] for discussion of Exploratory Data Analysis), making it possible to identify variables and trends of interest to follow up with more deductive methods.
+The same approach can be applied in most regions where there is a lack of information on which to build a robust provenance study.
+The main limitation is the material in question and how it responds to the spectroscopic instrumentation.
+The results are applicable to neighbouring regions and beyond, and should help in furthering research into prehistoric tools in Scandinavia and other areas.
 The material exhibits considerable variation in terms of texture and colour, and the results suggest that the initial classification, based on the archaeological documentation, may have included mistakes.
@@
 Ideally these different sources would be fused using methods such as multi-block PCA to better asses key variables contributing to the model [@mishra_etal21].
+Such analysis is, however, outside the scope of this paper and a potential direction for future work.
 
@@
 We would also like to acknowledge cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.
+Finally, we would like to thank the detailed comments and suggestions for improvement provided by two anonymous reviewers.
 
@@
 
-Johan Linderholm: Conceptualization, Methodology, Supervision
+Johan Linderholm: Conceptualization, Methodology, Supervision, Writing - Review & Editing
 

### 2024-09-24 README.Rmd
updated readme
fixing download zip link
Updated DOI in Readme

@@
 
-> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal of Archaeological Science:Reports) <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2024). _`r Title`_. (Journal of Archaeological Science:Reports) <https://doi.org/10.1016/j.jasrep.2024.104381>
 
@@
 
-> Sjölander M., Linderholm, J., Geladi P., Buckland, P. I. (2023). _`r Title`_. (Journal of Archaeological Science:Reports) `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P., Buckland, P.
+I.
+(2023).
+_`r Title`_.
+(Journal of Archaeological Science:Reports) `r format(Sys.Date(), "%d %b %Y")`.
+Available at SSRN: <http://dx.doi.org/10.2139/ssrn.4479650>
 
@@
 I.
-(2023).
+(2024).
 (`r format(Sys.Date(), "%Y")`).
@@
 Accessed `r format(Sys.Date(), "%d %b %Y")`.
-Online at <https://doi.org/xxx/xxx>
+Online at <https://doi.org/10.17605/OSF.IO/F6H98>
 
@@
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+[master.zip](https://github.com/MattiasSealander/quartzite.complexities/archive/refs/heads/main.zip). After unzipping:
 - open the `.Rproj` file in RStudio
